---
title: Arc length
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## Calculating arc length

Suppose we have a vector function $\vec{r}:[a,b] \to \mathbb{R}^n$:

$$\vec{r}(t) = \v{r_1(t), r_2(t), \dots}$$

where $a \le t \le b$.  We will assume that the components are
_differentiable_.

We want to find the _length_ of the curve from $\vec{r}(a)$ to $\vec{r}(b)$.

The idea is to first approximate the curve by a sequence of vectors like in
this illustration:

```{r}
illustr("../../asymptote/vector_functions/length.html", title = "Approximation of a curve by a sequence of vectors", width=600, height=400)
```

We then calculate the sum of the magnitudes of all the vectors, and let the
number of vectors go to infinity.  More details and some examples in the
following video:

```{r}
vembedr::embed_youtube("f1EuPPPoM3E")
```

### Try it yourself


```{r}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/13_Calculus_of_Vector-Valued_Functions/13.3_Arc_Length_and_Speed/13.3.3.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/Hope/Multi2/11-06-Scalar-line-integrals/Scalar-line-integral-01.pg",
        width=640, height=500
)
```
