---
title: Arc length
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Calculating arc length

A vector function $\vec{r}:[a,b] \to \mathbb{R}^n$:

$$\vec{r}(t) = \v{r_1(t), r_2(t), \dots}$$

where $a \le t \le b$.  We will assume that the components are
_differentiable_.

We want to find the _length_ of the curve from $\vec{r}(a)$ to $\vec{r}(b)$.

The idea is to first approximate the curve by a sequence of vectors like in
this illustration:

```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/length.html", title = "Approximation of a curve by a sequence of vectors", width=600, height=400)
```

We then calculate the sum of the magnitudes of all the vectors, and let the
number of vectors go to infinity.  More details and some examples in the
following video:

```{r echo=FALSE}
vembedr::embed_youtube("f1EuPPPoM3E")
```

### Try it yourself


```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/13_Calculus_of_Vector-Valued_Functions/13.3_Arc_Length_and_Speed/13.3.3.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Hope/Multi2/11-06-Scalar-line-integrals/Scalar-line-integral-01.pg",
        width=640, height=500
)
```
