---
title: Vector functions
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Vector functions

A vector function $\vec{r}:\mathbb{R} \to \mathbb{R}^n$:

$$\vec{r}(t) = \v{r_1(t), r_2(t), \dots}$$

There are several ways to interpret this.  Two of them are:

-   $\vec{r}(t)$ is a position vector of an object moving through the
  $n$-dimensional space at time $t$.

-   $\vec{r}$ is a "mapping" that takes points from the number line and places
  them into the $n$-dimensional space.  The point $t$ is placed at the location
  given by the position vector $\vec{r}(t)$.

If the function is _continuous_, it will look like this:

```{r echo=FALSE}
vembedr::embed_youtube("Q0WsJG_Uxik")
```

## Plotting vector functions and identifying the plots

As an example, consider the following vector function:

$$\vec{r}(t) = \v{1 + \cos 2\pi t, 1 + \sin 2\pi t, t}$$

Let's try to figure out how the curve defined by this function for $0 \le t \le 2$ looks like.

One useful tool for studying plots of vector functions is _projecting the curve
onto the coordinate planes_.

### Projection to the $xy$-plane

We have some curve in the 3d space.  Each point on this curve can be projected
to the $xy$-plane, simply by setting its $z$-coordinate to 0.  If we do that,
we get a new vector function 

$$\vec{r_{xy}}(t) = \v{1 + \cos 2\pi t, 1 + \sin 2\pi t, 0}$$

We know that in the $xy$-plane, the equations

$$\begin{aligned}
x &= \cos(2\pi t)\\
y &= \sin(2\pi t)
\end{aligned}$$

form a circle with center at the origin and radius 1, traveled
counterclockwise, starting at $(1,0)$ for $t = 0$, and going once around the
whole circle every time $t$ increases by 1. This follows from the very
definition of _trigonometric functions_ $\cos$ and $\sin$.

Here we have 

$$\begin{aligned}
x &= 1 + \cos(2\pi t)\\
y &= 1 + \sin(2\pi t)
\end{aligned}$$

which will move the circle so that its center is at the point $(1,1)$.

So the projection of our curve onto the $xy$-plane is the circle with radius 1,
center at $(1,1)$, traveled counterclokwise _twice_ around, starting at the
point $(2,1)$.


```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/projectionxy.html", title = "Projection onto the xy-plane", width=400, height=400)
```

### Projection to the $xz$-plane

To project the points on the curve to the $xz$-plane, we set its $y$-coordinate to 0.
This will give us a vector function 

$$\vec{r_{xz}}(t) = \v{1 + \cos 2\pi t, 0, t}$$

In other words, in the $xz$-plane, we have equations

$$\begin{aligned}
x &= 1 + \cos(2\pi t)\\
z &= t
\end{aligned}$$

The second equation says that $t = z$.  We can plug this into the first
equation, eliminating the $t$, and getting

$$x = 1 + \cos(2\pi z) \text{ for } 0 \le z \le 2$$.

Here we have $x$ as a function of $z$, a function which we can graph:

```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/projectionxz.html", title = "Projection onto the xz-plane", width=400, height=400)
```

### Projection to the $yz$-plane

In a silimar way we can see that the projection onto the $yz$-plane will be the
graph of the function $y = 1 + \sin(2\pi z)$ for $0 \le z \le 2$:

```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/projectionyz.html", title = "Projection onto the yz-plane", width=400, height=400)
```

### The graph of the vector function with the projections

```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/projections.html", title = "Graph of the vector function with projections", width=2*285, height=2*256)
```

### More examples

```{r echo=FALSE}
vembedr::embed_youtube("ObGDHSH8XP0")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart13_1/problem_3.pg",
        width=640, height=500
)
```

## Domain of a vector function

In order for $\vec{r}(t)$ to be defined, __all of the components__ $r_1(t)$,
$r_2(t)$, ... $r_n(t)$ must be defined.  For example:

Consider the vector function $\vec{r}(t) = \v{1/t, \sqrt{t+1}, \ln(5-t)}$.

-  The first component is undefined at $t = 0$.
-  The second component is undefined at $t < -1$.
-  The third component is undefined at $t \ge 5$.

The domain of the function is then the union of two intervals: $[-1,0) \cup (0,5)$.

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Dartmouth/setStewartCh14S1/problem_1.pg",
        width=640, height=500
)
```

## Limits and continuity

Intuitively, a continuous vector function would correspond to a trajectory of a
moving object where the object does not do things like sudden jumps through the
space, does not cease to exist for a moment, does not fly off to an infinite
distance only to come back from an infinite distance, whether form the same or
different direction, and so on.

### Examples

1.  The function

    $$\vec{r}(t) = \v{t^2, 2t, \frac{1}{t-1}}$$

    is discontinuous at $t = 1$.  As $t$ approaches 1 from the left, the
    $x$-coordinate approaches 1, the $y$-coordinate approaches 2, while the
    $z$-coordinate goes off towards $-\infty$.  As $t$ approaches 1 from the right,
    the $x$ and $y$-coordinates still approach 1 and 2, respectively, but the
    $z$-coordinate now goes off to $\infty$.

    On the illustration below, the part of the curve for $t < 1$ is shown in red,
    while the part for $t>1$ is blue.  The arrows show the direction of increasing
    $t$.  The view is limited to a finite part of the space, in reality the curve
    stretches indefinitely down and up.

    ```{r, echo=FALSE}
    illustr("../../asymptote/vector_functions/discontinuity1.html", title = "Example of a discontinuous curve with a linear asymptote", width=285, height=256)
    ```

2.  The function

    $$\vec{r}(t) = \v{2t, \frac{t+1}{(t-1)^2}, \frac{1}{t-1}}$$

    is discontinuous at $t = 1$.  As $t$ approaches 1 from the left, the
    $x$-coordinate approaches 1, and the $y$-coordinate goes off towards $\infty$
    and the $z$-coordinate goes off towards $-\infty$.  As $t$ approaches 1 from the right,
    the $x$ still approaches 1, the $y$-coordinate goes off to $\infty$, and the
    $z$-coordinate now also goes off to $\infty$.

    On the illustration below, the part of the curve for $t < 1$ is shown in red,
    while the part for $t>1$ is blue.  The arrows show the direction of increasing
    $t$.  The view is limited to a finite part of the space, in reality the curve
    stretches indefinitely down and right and up and right.

    ```{r, echo=FALSE}
    illustr("../../asymptote/vector_functions/discontinuity2.html", title = "Example of a discontinuous curve with a parabolic asymptote", width=285, height=256)
    ```

3.  The function

    $$\vec{r}(t) = \v{\frac{t-1}{\left\lvert t-1\right\rvert} + 2t, t^2 - 1, \frac{1}{t*2+1} - \frac{1}{2}}$$

    is discontinuous at $t = 1$.  As $t$ approaches 1 from the left, the
    $x$-coordinate approaches 1, the $y$-coordinate approached 0
    and the $z$-coordinate approaches 0.  As $t$ approaches 1 from the right,
    the $x$ approaches 3, and the $y$-coordinate and the $z$-coordinate still
    both approach 0.

    On the illustration below, the part of the curve for $t < 1$ is shown in red,
    while the part for $t>1$ is blue.  The arrows show the direction of increasing
    $t$.

    ```{r, echo=FALSE}
    illustr("../../asymptote/vector_functions/discontinuity3.html", title = "Example of a discontinuous curve with a jump discontinuity", width=285, height=256)
    ```

There are many other things that can happen, these are just some simple
examples that are easy to visualize.

### Limits

```{r echo=FALSE}
vembedr::embed_youtube("JFRKL5I8Vgw")
```

#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/13_Calculus_of_Vector-Valued_Functions/13.2_Calculus_of_Vector-Valued_Functions/13.2.1.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart13_1/problem_2.pg",
        width=640, height=500
)
```

