---
title: TNB Frame, Osculating Planes and Circles
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## TNB Frame

At ant point of a regularly parametrized curve, the unit _tangent_ vector, the
unit _normal_ vector and the unit _binormal_ vector form a local coordinate
frame called the TNB frame:

```{r echo=FALSE}
vembedr::embed_youtube("dvOl7bUy9ns")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart13_3/problem_4.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Rochester/setVecFunction2Curvature/ur_vc_4_4.pg",
        width=640, height=500
)
```

## Osculating plane and circle

We already know that a regularly parametrized curve can be locally approximated
by a line called the _tangent line_.  At points where the curvature is not
zero, we can get even better approximation by a circle.  The circle that
provides the _best approximation_ of the curve at a given point is called the
_osculating circle_ of the curve at that point:

```{r echo=FALSE}
vembedr::embed_youtube("0lar2hAz_zY")
```

## An example of TNB frame and osculating circle:

The following video shows an animation which visualizes a TNB frame and
osculating circle moving along a curve:

```{r echo=FALSE}
vembedr::embed_youtube("JZGFcwipHYY")
```

The following illustration shows the curve from the above video with several of the TNB frames:

```{r, echo=FALSE}
illustr("../../asymptote/vector_functions/tnb_pic.html", title = "A curve with several TNB frames", width=600, height=400)
```

