---
title: TNB Frame, Osculating Planes and Circles
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## TNB Frame

At any point of a regularly parametrized curve, the unit _tangent_ vector, the
unit _normal_ vector and the unit _binormal_ vector form a local coordinate
frame called the TNB frame:

```{r}
vembedr::embed_youtube("dvOl7bUy9ns")
```

### Try it yourself

```{r}
webwork(
        "Library/272/setStewart13_3/problem_4.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/Rochester/setVecFunction2Curvature/ur_vc_4_4.pg",
        width=640, height=500
)
```

## Osculating plane and circle

We already know that a regularly parametrized curve can be locally approximated
by a line called the _tangent line_.  At points where the curvature is not
zero, we can get even better approximation by a circle.  The circle that
provides the _best approximation_ of the curve at a given point is called the
_osculating circle_ of the curve at that point:

```{r}
vembedr::embed_youtube("0lar2hAz_zY")
```

## An example of TNB frame and osculating circle:

The following video shows an animation which visualizes a TNB frame and
osculating circle moving along a curve:

```{r}
vembedr::embed_youtube("JZGFcwipHYY")
```

The following illustration shows the curve from the above video with several of the TNB frames:

```{r}
illustr("../../asymptote/vector_functions/tnb_pic.html", title = "A curve with several TNB frames", width=600, height=400)
```

