---
title: Velocity and Acceleration
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## Velocity and acceleration

As we already saw, a vector function $\vec{r}:[a,b] \to \mathbb{R}^3$:

$$\vec{r}(t) = \v{x1(t), y(t), z(t)}$$

for $a \le t \le b$ can be interpreted as a position vector of a moving point
in $\mathbb{R}^3$ at the time $t$.  As the point moves through the space, it
traces a trajectory, which is a curve in $\mathbb{R}^3$.  We say that the
vector function $\vec{r}$ is a __parametrization__ of the curve.

We also already know that the derivative $\vec{r}'(t)$ can be interpreted as
the _velocity vector_ of the moving point -- it will tell us how fast and in
which direction is the position changing. The magnitude of the velocity vector
$\norm{\vec{r}'(t)}$ is the _speed_ of the motion.

The second derivative $\vec{r}''(t)$ will tell us how the velocity changes.  It
can be interpreted as the _acceleration vector_.

### Try it yourself

```{r}
webwork(
        "Library/272/setStewart13_4/problem_7.pg",
        width=640, height=500
)
```

## Multiple parametrizations

Different vector functions can correspond to the same trajectory in
$\mathbb{R}^3$. We can think about it as two objects traveling along the same
curve, but at different times and at possibly different speeds (or even in the
opposite direction).

We will only consider so called _regular_ parametrizations, which are vector
functions that are continuous on $[a,b]$, _twice differentiable_ on $(a,b)$, 
and $\norm{\vec{r}'(t)} \neq 0$ for all $t\in (a,b)$.

The following video explores relations between two regular parametrizations of
the same curve.

```{r}
vembedr::embed_youtube("NSf84jrJP0I")
```

## Tangential and normal components of acceleration

When the velocity of a motion change, there are two things that can be
changing:

-  the speed, or magnitude of the velocity

-  the direction of the motion

The following video looks at how an acceleration vector can cause these two
changes.

```{r}
vembedr::embed_youtube("K3iWUK2Wd_o")
```

### Try it yourself

```{r}
webwork(
        "Library/Dartmouth/setStewartCh14S2/problem_3.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/ASU-topics/setCalculus/stef/stef13_4p2.pg",
        width=640, height=500
)
```

## Curvature

In the previous video we saw that the curvature is defined as

$$\kappa(t) = \frac{\norm{\text{normal component of } \vec{r}''(t)}}{\norm{\vec{r}'(t)}^2}$$

and can be calculated as 

$$\kappa(t) = \frac{\norm{\vec{T}'(t)}}{\norm{\vec{r}'(t)}}$$

Most of the time, calculating the derivative of the unit tangent vector is
hard.  In this video we will derive a formula for curvature that is usually
easier to use:

```{r}
vembedr::embed_youtube("12ZG8j9f0Oo")
```

### Try it yourself


```{r}
webwork(
        "Library/272/setStewart13_3/problem_6.pg",
        width=640, height=500
)
```

