---
title: Changing the Order of Integration
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../../_common.qmd >}}

## Fubini's Theorem

When discussing double integrals and iterated integrals, we learned about
the following theorem:

__Theorem:__ Let $f$ be a function of two variables defined on a region $R$
such that $f$ and $R$ are nice.
Then the double integral of $f$ over the region $R$ and the two iterated
integrals over the same region are equal:

$$\iint_R f(x,y)\;dA = \int_c^d \int_{a(y)}^{b(y)} f(x,y)\;dx\;dy = \int_a^b \int_{c(x)}^{d(x)} f(x,y)\;dy\;dx$$

So far we have used this theorem to _evaluate double integrals_ by turning them
into iterated integrals. We have also seen that for some regions $R$, we can
rewrite the double integrals as an iterated integral two ways, both $dx\;dy$
and $dy\;dx$. We also saw that depending on the function $f$ and the region
$R$, _one of these two iterated integrals is often much easier to calculate_.

That can be used to calculate some difficult iterated integrals. Suppose you
have an iterated integral that is hard to calculate. It is possible that by
rewriting the iterated integral as a _double integral_, and then rewriting the
double integral as an iterated integral with the opposite order of integration
from the original iterated integral, the resulting iterated integral could be
easier to calculate.

## Examples of changing order of integration

Let's look at some examples:

### Example 1

Calculate $$\int_0^3\int_{x^2/3}^x xy\;dy\;dx$$

```{r}
vembedr::embed_youtube("kzs4vrOpF78")
```

### Example 2

Calculate $$\int_0^{2\sqrt{\pi}}\int_{y/2}^{\sqrt{\pi}} \cos\left(x^2\right)\;dx\;dy$$

```{r}
vembedr::embed_youtube("uRwPxXzD4Rk")
```

### Try it yourself

```{r}
webwork(
        "Library/FortLewis/Calc3/16-2-Iterated-integrals/HGM4-16-2-38-Iterated-integrals.pg",
        width=640, height=600
)
```

```{r}
webwork(
        "Library/Michigan/Chap16Sec2/Q35.pg",
        width=640, height=500
)
```
