---
title: Substitution for Multiple Integrals
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Substitution (change of variable) in multiple integrals

### Explanation and derivation:

```{r echo=FALSE}
vembedr::embed_youtube("KlNcZcI4wpc")
```

### Summary, Jacobians

```{r echo=FALSE}
vembedr::embed_youtube("CTsRERGu_H4")
```

#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap16Sec7/Q01.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Hope/Multi2/11-04-MV-Int-change-var/Change-variables-01.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/15_Multiple_Integration/15.6_Change_of_Variables/15.6.23.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap16Sec7/Q05.pg",
        width=640, height=500
)
```

### Examples

#### Polar coordinate transformation

```{r echo=FALSE}
vembedr::embed_youtube("3nTbXYz6HWA")
```

#### Another example of substitution in 2 variables

```{r echo=FALSE}
vembedr::embed_youtube("lNi6s0b6HKc")
```

#### An example of substitution in 3 variables (spherical coordinate transformation)

```{r echo=FALSE}
vembedr::embed_youtube("rOsUNGU2H9s")
```

#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap16Sec7/Q19.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/15_Multiple_Integration/15.6_Change_of_Variables/15.6.33.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap16Sec7/Q25.pg",
        width=640, height=500
)
```
