---
title: "Green's Theorem"
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
    warning: false
---

{{< include ../_common.qmd >}}

```{r, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
```

## Another test for conservativeness

Given a differentiable function $f\colon \mathbb{R}^2 \to \mathbb{R}$, its
_gradient_ is a _conservative_ vector field
$$\vec{F}(x,y) = \v{P(x,y), Q(x,y)} = \vec{\nabla}f(x,y) = \v{f_x(x,y),f_y(x,y)}$$

According to _Clairaut's Theorem_, if $f$ has _continuous second partial
derivatives_, then the _mixed second partial derivatives_ $f_{xy}$ and $f_{yx}$
are equal.

Since for a conservative vector field $\vec{F} = \v{P,Q}$, the components are
$P = f_x$ and $Q = f_y$, the _second partial derivatives_ of $f$ are the _first
partial derivatives_ of $P$ and $Q$, and the _mixed second partial derivatives_ of $f$ are

\begin{aligned}
f_{xy} &= P_y\\
f_{yx} &= Q_x
\end{aligned}

In other words, if we have a vector field $\vec{F} = \v{P, Q}$ such that both
$P$ and $Q$ have _continuous partial derivatives_, and $P_y \neq Q_x$, the
field $\vec{F}$ __cannot be conservative__! That gives us a new easier way to
detect _non-conservative_ vector fields without having to attempt to find a
potential function!

```{r echo=FALSE}
vembedr::embed_youtube("QBvLyMI_h70")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/ASU-topics/setCalculus/stef16_3p1.pg",
        width=640, height=500
)
```


### Some questions

There are some obvious questions we can ask about this:

1.  What's actually going on?  What does the difference between $P_y$ and $Q_x$
    have to do with conservativeness of the vector field?

2.  Does it work the other way? Can we say that if $P_y(x,y) = Q_x(x,y)$ then
    the vector field is conservative?

3.  Can we extend this to 3 variables?

Let's first graph the two examples of non-conservative vector fields, maybe it
will give us some idea of what is going on:

*   The field $\vec{F}(x,y) = \v{-y, x}$:

    ```{r}
    expand.grid(x = seq(-1,1,length.out=20), y = seq(-1,1,length.out=20)) %>%
        mutate(dx = -y, dy = x) %>%
        ggplot(aes(x,y)) + geom_segment(aes(xend = x+0.08*dx, yend =
        y+0.08*dy), arrow = arrow(length=unit(0.1,"cm"))) + coord_fixed() +
        xlim(-1,1) + ylim(-1,1)
    ```

    We can see clearly that this is not a conservative vector field: any
    integral over a circle centered at $(0,0)$ is going to be positive! This
    field seems to represent some sort of _rotation_ around the origin.

*   The field $\vec{F}(x,y) = \v{e^x\cos y, e^x\sin y}$:

    ```{r}
    expand.grid(x = seq(-1.5,1.5,length.out=20), y = seq(0,3,length.out=20)) %>%
        mutate(dx = exp(x)*cos(y), dy = exp(x)*sin(y)) %>%
        ggplot(aes(x,y)) + geom_segment(aes(xend = x+0.08*dx, yend =
        y+0.08*dy), arrow = arrow(length=unit(0.1,"cm"))) + coord_fixed() +
        xlim(-1.5,1.5) + ylim(0,3)
    ```

    Again, we can see why this is non-conservative: while most arrows point
    generally upwards, the arrows on the right are much longer than the arrows
    on the left. If you insert any object into this force field somewhere near
    the center of the plot, the upward forces on the right side of the object
    will be much bigger than the upward forces on the left side of the object,
    causing the object to rotate counter-clockwise. An integral over any circle
    with center near the center of the plot window will be positive.

We want to better understand what does the equality of the partial derivatives
of the field components have to do with conservativeness. Since derivatives are
"local" properties, it may help to look at things at a really tiny scale. Let's
calculate a line integral of a vector field over a very very tiny rectangle
centered at $(x_0,y_0)$, with sides of length $dx$ and $dy$. We will label the four
sides of the rectangle as $S_1$, $S_2$, $S_3$ and $S_4$:

```{r,engine='tikz',fig.ext='svg',fig.width=5, engine.opts=list(dvisvgm.opts = "--font-format=woff")}
\usetikzlibrary{decorations.markings}
\begin{tikzpicture}
\fill (1.5, 1.5) circle[radius=1pt] node[below]{$(x_0,y_0)$};
\begin{scope}[thick,decoration={
    markings,
    mark=at position 0.7 with {\arrow{>}}}
    ] 
    \draw[postaction={decorate}] (0,0) --node[below]{$S_1$} (3,0);
    \draw[postaction={decorate}] (3,0) --node[right]{$S_2$} (3,3);
    \draw[postaction={decorate}] (3,3) --node[above]{$S_3$} (0,3);
    \draw[postaction={decorate}] (0,3) --node[left]{$S_4$} (0,0);
\end{scope}
\fill (0,0) circle[radius=1pt] node[below left]{$(x_0-\frac{dx}{2},y_0-\frac{dy}{2})$};
\fill (3,0) circle[radius=1pt] node[below right]{$(x_0+\frac{dx}{2},y_0-\frac{dy}{2})$};
\fill (3,3) circle[radius=1pt] node[above right]{$(x_0+\frac{dx}{2},y_0+\frac{dy}{2})$};
\fill (0,3) circle[radius=1pt] node[above left]{$(x_0-\frac{dx}{2},y_0+\frac{dy}{2})$};
\end{tikzpicture}
```

Then the integral of $\vec{F}$ over the rectangle can be split into the sum of
the integrals over the four sides:

$$ \int_{S_1} \vec{F}(x,y)\cdot \vec{dr} + \int_{S_2} \vec{F}(x,y)\cdot \vec{dr} + \int_{S_3} \vec{F}(x,y)\cdot \vec{dr} + \int_{S_4} \vec{F}(x,y)\cdot \vec{dr} $$

Reversing the direction of the integrals over $S_3$ and $S_4$ will give us

$$ \int_{S_1} \vec{F}(x,y)\cdot \vec{dr} + \int_{S_2} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_3} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_4} \vec{F}(x,y)\cdot \vec{dr} $$

Let's approximate the four integrals:

*   We can parametrize the $S_1$ curve as $x = x_0 + t$, 
    $y = y_0 - \frac{dy}{2}$ for $-\frac{dx}{2} \le t \le \frac{dx}{2}$.  
    Then $dx = dt$ and $dy = 0$, and so
    $$\int_{S_1} \vec{F}(x,y)\cdot\vec{dr} = \int_{-\frac{dx}{2}}^{\frac{dx}{2}} P\left(x_0 + t, y_0 - \frac{dy}{2}\right)\;dt$$

    If $dx$ is very small, the last integral is approximately equal to 
    $$P\left(x_0, y_0 - \frac{dy}{2}\right)\;dx$$

*   We can parametrize the $-S_3$ curve as $x = x_0 + t$, 
    $y = y_0 + \frac{dy}{2}$ for $-\frac{dx}{2} \le t \le \frac{dx}{2}$.  
    Then $dx = dt$ and $dy = 0$, and so
    $$\int_{-S_3} \vec{F}(x,y)\cdot\vec{dr} = \int_{-\frac{dx}{2}}^{\frac{dx}{2}} P\left(x_0 + t, y_0 + \frac{dy}{2}\right)\;dt$$

    If $dx$ is very small, the last integral is approximately equal to 
    $$P\left(x_0, y_0 + \frac{dy}{2}\right)\;dx$$

*   We can parametrize the $S_2$ curve as $x = x_0 + \frac{dx}{2}$, 
    $y = y_0 + t$ for $-\frac{dy}{2} \le t \le \frac{dy}{2}$.  
    Then $dx = 0$ and $dy = dt$, and so
    $$\int_{S_2} \vec{F}(x,y)\cdot\vec{dr} = \int_{-\frac{dy}{2}}^{\frac{dy}{2}} Q\left(x_0 + \frac{dx}{2}, y_0 + t\right)\;dt$$

    If $dy$ is very small, the last integral is approximately equal to 
    $$Q\left(x_0 + \frac{dy}{2}, y_0\right)\;dy$$

*   We can parametrize the $-S_4$ curve as $x = x_0 - \frac{dx}{2}$, 
    $y = y_0 + t$ for $-\frac{dy}{2} \le t \le \frac{dy}{2}$.  
    Then $dx = 0$ and $dy = dt$, and so
    $$\int_{S_2} \vec{F}(x,y)\cdot\vec{dr} = \int_{-\frac{dy}{2}}^{\frac{dy}{2}} Q\left(x_0 - \frac{dx}{2}, y_0 + t\right)\;dt$$

    If $dy$ is very small, the last integral is approximately equal to 
    $$Q\left(x_0 - \frac{dy}{2}, y_0\right)\;dy$$

Looking at the two _horizontal sides_: if $dx$ is very small,

\begin{aligned}
\int_{S_1} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_3} \vec{F}(x,y)\cdot \vec{dr}
&\approx P\left(x_0, y_0 - \frac{dy}{2}\right)\;dx - P\left(x_0, y_0 + \frac{dy}{2}\right)\;dx \\
&= \left(P\left(x_0, y_0 - \frac{dy}{2}\right) - P\left(x_0, y_0 + \frac{dy}{2}\right)\right)\;dx \\
&= \left(\frac{P\left(x_0, y_0 - \frac{dy}{2}\right) - P\left(x_0, y_0 + \frac{dy}{2}\right)}{dy}\right)\;dx\;dy
\end{aligned}

When $dy$ is very small, 
$$\frac{P\left(x_0, y_0 - \frac{dy}{2}\right) - P\left(x_0, y_0 + \frac{dy}{2}\right)}{dy} \approx -P_y(x_0, y_0)$$

If both $dx$ and $dy$ are very small, then
$$\int_{S_1} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_3} \vec{F}(x,y)\cdot \vec{dr} \approx -P_y(x_0, y_0)\;dx\;dy$$

Let's now look at the two _vertical sides_: if $dy$ is very small,

\begin{aligned}
\int_{S_2} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_4} \vec{F}(x,y)\cdot \vec{dr}
&\approx Q\left(x_0 + \frac{dx}{2}, y_0\right)\;dy - Q\left(x_0 + \frac{dx}{2}\right)\;dy \\
&= \left(Q\left(x_0 + \frac{dx}{2}, y_0\right) - Q\left(x_0 - \frac{dx}{2}, y_0\right)\right)\;dy \\
&= \left(\frac{Q\left(x_0 + \frac{dx}{2}, y_0\right) - Q\left(x_0 - \frac{dx}{2}, y_0\right)}{dx}\right)\;dx\;dy
\end{aligned}

When $dx$ is very small, 
$$\frac{Q\left(x_0 + \frac{dx}{2}, y_0\right) - Q\left(x_0 - \frac{dx}{2}\right)}{dx} \approx Q_x(x_0, y_0)$$

If both $dx$ and $dy$ are very small, then
$$\int_{S_2} \vec{F}(x,y)\cdot \vec{dr} - \int_{-S_4} \vec{F}(x,y)\cdot \vec{dr} \approx Q_x(x_0, y_0)\;dx\;dy$$

Putting all this together: when both $dx$ and $dy$ are very small, then the
integral of $\vec{F}(x,y)$ over the boundary of the small rectangle centered at
$(x_0, y_0)$ with sides $dx$ and $dy$ is approximately equal to
$$\left(Q_x(x_0, y_0) - P_y(x_0, y_0)\right)\;dx\;dy$$
This approximation gets better the smaller $dx$ and $dy$ are^[This argument can
be actually made precise, without the need to refer to approximations, by using
the Mean Value Theorem for derivatives and the Mean Value Theorem for definite
integrals. It makes the whole derivation much more complicated, though.].

## Simple Curves and Circulation

```{r echo=FALSE}
vembedr::embed_youtube("9sJ8YGSvU60")
```

## Green's Theorem

We saw that the quantity $$\left(Q_x(x,y) - P_y(x,y)\right)\;dx\;dy$$
can be interpreted as a "local circulation" over an infinitesimal rectangle
centered at $(x,y)$ with sides $dx$ and $dy$. We also was that circulation is
"additive" in the sense that circulation over two simple curves that are
adjacent to each other can be added. Intuitively this leads one to think that
perhaps "summing" all local circulations at all interior points of a simple
curve should give us a number that is equal to the circulation over the curve
itself. Formally, this is expressed as the __Green's Theorem__^[The form of
Green's Theorem presented here is known as the _Circulation form_ or _Tangential
form_. There is also a form known as _Flux form_ or _Normal form_. We are not
covering it in this class, but you can find it in your textbook.]:

_Suppose $G$ is a region in a plane whose boundary $\partial G$ is a piecewise
smooth simple closed curve. Let $\vec{F}(x,y) = \v{P(x,y), Q(x,y)}$ be a vector
field defined on $G$ whose components have continuous partial derivatives on
$G$. Then_
$$\oint_{\partial G} \vec{F}(x,y)\cdot\vec{dr} = \iint_G \left(Q_x(x,y) - P_y(x,y)\right)\;dA$$

Above we gave an intuitive justification for the Green's Theorem. The formal
proof is beyond the scope of this class. 

Note that this theorem is really a two-dimensional form of the Fundamental
Theorem of Calculus. It says that an integral of some specific type of
_derivative_ of a vector field $\vec{F}$ over a region $G$ is equal to the
integral of the vector field itself over the _boundary_ of the region $G$.

In the one-dimensional case, the "vector field" is just a real function of one
variable, the "region" is just an interval, and its boundary consists of just
two points.

### Applications of Green's Theorem

```{r echo=FALSE}
vembedr::embed_youtube("IA3UINSjWJg")
```

One more important consequence of Green's Theorem is that the answer to our
question 2 above is affirmative: If $\vec{F} = \v{P(x,y), Q(x,y)}$ is a vector field whose
components have continuous partial derivatives, then $\vec{F}$ is conservative
_if and only if_ $Q_x(x,y) = P_y(x,y)$.

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap18Sec4/Q15.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/17_Fundamental_Theorems_of_Vector_Analysis/17.1_Greens_Theorem/17.1.3.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart16_4/ur_vc_12_8.pg",
        width=640, height=500
)
```
