---
title: "\"Line\" Integrals of Vector Fields"
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## Vector Fields

So far we have discussed the following types of functions:

In Calc 1 (actualy in an Algebra class or a Pre-calculaus class) you learned
about real functions of a single variable.  These take one real number as input
(_single_ variable) and return a single real number (_real_ function).
Symbolically, this is written as $f\colon \mathbb{R} \to \mathbb{R}$.

Earlier in Calc 3, we have learned about _vector_ functions of a single
variable: these take a single real number as input and return a vector as
output. Symbolically we can write this as $\vec{r}\colon \mathbb{R} \to
\mathbb{R}^n$.

Also in Calc 3, we saw real functions of several variables. These take a point
or vector as input, and return a single real number. Symbolically,
$f\colon\mathbb{R}^n \to \mathbb{R}$.

Now we are going to combine the last two ideas together and start looking at
_vector_ functions of _several variables_. For start, we will look at one
simpler case where the dimension of the input is the same as the dimension of
the output: functions from $\mathbb{R}^n$ to $\mathbb{R}^n$.

A function from $\mathbb{R}^n$ to $\mathbb{R}^n$ can be interpreted in
different ways, some of which we will see later. For now, we will think about
such functions as __vector fields__: the input is interpreted as a __point__ in
$\mathbb{R}^n$, while the output is a __vector__.  In other words, each point
is the domain of the function has a vector associated with it.

### Examples

Some examples of phenomena that can be modeled as vector fields:

*  __Flow of a fluid__: We saw that a vector function of one variable can be
    used to model a motion of a single object in a plane or space. In similar
    way we can model a flow of a liquid or gas on a plane or in a space: at
    each point, the flowing fluid has a velocity that determines the speed and
    direction of the flow.  This velocity can be described by a vector. If the
    flow is constant in time, this will give us a vector field.  If the flow
    changes with time, there will be an additional variable that represents
    time, and instead of a single vector field, we will get a parametrized
    family of vector fields. For each specific value of the time parameter, we
    would have a single vector field.

*  __A force field__: In the past we used vectors to model forces acting on
    objects. Vector fields can be used to model situations where there is a
    force at every point of a plane or space, like gravitational or
    electromagnetic force.

*   __Gradient of a function__: we have already seen one example of a vector
    field: the _gradient vector_ of a function of several variables. At every
    point of the domain of $f$, the vector $\vec{\nabla} f$ points in the
    direction of the steepest increase of $f$, and its magnitude gives us the
    rate of change of $f$ in that direction.

### Components of a vector field

A vector field $\vec{F}: \mathbb{R}^3 \to \mathbb{R}^3$ gives us a vector for
each point $(x,y,z)$ in the domain of $\vec{F}$. This vector can be written in
components: $$\vec{F} = \v{P, Q, R} = P\vec{i} + Q\vec{j} + R\vec{k}.$$
The components $P$, $Q$ and $R$ _depend on the point $(x,y,z)$_, and are
therefore functions from $\mathbb{R}^3$ to $\mathbb{R}$, in other words, _real
functions of three variables_. We can write it this way:
$$\vec{F}(x,y,z) = \v{P(x,y,z), Q(x,y,z), R(x,y,z)} = P(x,y,z)\vec{i} + Q(x,y,z)\vec{j} + R(x,y,z)\vec{k}$$

The functions $P$, $Q$ and $R$ are called the _component functions_ of the
field $\vec{F}$.

### Differentiation of vector fields

Just like with vector functions, limits and therefore derivatives of vector
fields are done by taking limits and derivatives of the component functions.
Since vector fields and their component functions are functions of several
variables, the derivatives will be _directional derivatives_ and, in
particular, _partial derivatives_. When all component function of a vector
field are _differentiable_ at a point, we say that the vector field is
_differentiable_ at the point.

#### Example:

Find the partial derivatives of the vector field $$\vec{F}(x,y) = \v{x^4y^3,
x^2 + y^5}$$

Here the _component functions_ are $P(x,y) = x^4y^3$ and $Q(x,y) = x^2 + y^5$.

*  Partial derivative with respect to $x$: 
    $$\vec{F}_x(x,y) = \v{P_x(x,y), Q_x(x,y)} = \v{4x^3y^3, 2x}$$
*  Partial derivative with respect to $y$: 
    $$\vec{F}_y(x,y) = \v{P_y(x,y), Q_y(x,y)} = \v{3x^4y^2, 5y^4}$$

Later we will learn about some more interesting _differential operators_ on
vector fields.

## Line integrals of vector fields

```{r echo=FALSE}
vembedr::embed_youtube("yjNpDaENU7g")
```


### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/17-3-Vector-fields/HGM4-17-3-16a-Vector-fields.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/17-3-Vector-fields/HGM5-17-3-27-Vector-fields.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/17-3-Vector-fields/HGM4-17-3-00-Vector-fields.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/16_Line_and_Surface_Integrals/16.1_Vector_Fields/16.1.3.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/18-1-Idea-of-line-integrals/HGM4-18-1-02-Line-integrals.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/UBC/calculusStewart/lineIntegrals21.pg",
        width=640, height=500
)
```
