---
title: 3d Coordinate System
format: html
knitr:
  opts_chunk: 
    echo: false
---

```{r}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

### Introduction

```{r}
vembedr::embed_youtube("JsLqgb1E2c0")
```

### Coordinate planes

The following illustration shows the 3d coordinate system with the three
_coordinate planes_:

*  The $xy$-plane, which contains the $x$-axis and the $y$-axis, and is perpendicular to the $z$-axis

   The points on this plane have $z$-coordinate equal to 0.

*  The $xz$-plane, which contains the $x$-axis and the $z$-axis, and is perpendicular to the $y$-axis

   The points on this plane have $y$-coordinate equal to 0.

*  The $yz$-plane, which contains the $y$-axis and the $z$-axis, and is perpendicular to the $x$-axis

   The points on this plane have $x$-coordinate equal to 0.

```{r}
illustr("../../asymptote/3dcoords/coordplanes.html", title = "Three coordinate planes", width=285*2, height=256*2)
```

You can interact with the illustration by turning it around and viewing it
from different angles.

### __Projections__ of a point

A point $P$ with coordinates $(x,y,z)$ in a 3d coordinate system has a number of _projections_.

- First there are the projections to the _coordinate planes_:

    - Its projection to the $xy$ plane is the __nearest__ point to $P$ that lies on
      the $xy$-plane.  In other words, this is the nearest point to $P$ with zero
      $z$-coordinate.  Its coordinates are $(x,y,0)$.

    - Its projection to the $xz$ plane is the __nearest__ point to $P$ that lies on
      the $xz$-plane.  In other words, this is the nearest point to $P$ with zero
      $y$-coordinate.  Its coordinates are $(x,0,z)$.

    - Its projection to the $yz$ plane is the __nearest__ point to $P$ that lies on
      the $yz$-plane.  In other words, this is the nearest point to $P$ with zero
      $x$-coordinate.  Its coordinates are $(0,y,z)$.

- Then there are projections of $P$ to each of the three _coordinate axes_:

    - The projection to the $x$-axis is the point on the $x$-axis that is
      __nearest__ to $P$.  Since it is on the $x$-axis, the $y$ and $z$
      coordinates must both be 0.  The coordinates of this projection are
      $(x,0,0)$.

    - Similarly, the projection of $P$ to the $y$-axis, or the point on the
      $y$-axis that is __nearest__ to $P$, is $(0,y,0)$.

    - Finally, the projection of $P$ to the $z$-axis, or the point on the
      $z$-axis that is __nearest__ to $P$, is $(0,0,z)$.

Those 6 projections, together with the point $P$ itself and the _origin_, form a
rectangular box with the point $P$ and the _origin_ at the directly opposite
corners:

```{r}
illustr("../../asymptote/3dcoords/coordbox.html", title = "Projections and coordinate box", width=285*2, height=256*2)
```

If $P$ lies on one of the coordinate planes, axes, or at the origin, this box
will _degenerate_ into a rectangle, an interval, or a single point.

The following illustration combines both the "coordinate box" and the
coordinate planes, so you can see the projections on the actual planes:

```{r}
illustr("../../asymptote/3dcoords/coordplanesbox.html", title = "Coordinate box with coordinate planes", width=285*2, height=256*2)
```

### Try it yourself

```{r}
webwork("Library/272/setStewart12_1/problem_1.pg", width=640, height=500)
```

### A bit of a stretch

See if you can figure out the answers to these (_Hint:_ the distance from a point
$P$ to a plane or a line is a distance from $P$ to the point on the plane or
line that is __nearest__ to $P$.)

```{r}
webwork("Library/UMN/calculusStewartET/s_12_1_10.pg", width=640, height=500)
```

