---
title: Distance in 3d
format: html
knitr:
  opts_chunk: 
    echo: false
---

```{r}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

### Distance in 2d

We all know the _distance formula_ in plane.  It is based on the Pythagorean
Theorem.  If you draw a rectangle whose opposite corners are the two points
$(x_1,y_1)$ and $(x_2,y_2)$ and whose sides are parallel to the coordinate
axes, the distance between the two points is the length of the diagonal:

```{r}
illustr("../../asymptote/distance/2ddistance.html", title = "Rectangle with corners at the two given points", width=385, height=256)
```

Applying the Pythagorean Theorem on one of the right triangles in the picture
will tell us that the distance between $(x_1, y_1)$ and $(x_2,y_2)$ is
$$\sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$$

:::{.callout-tip}
#### A study tip

In order to fully understand a mathematical idea such as a definition, a
theorem or a formula, it is often useful to think about _special cases_.
In the distance formula, what would happen if the two points have the same
$x$-coordinate or $y$-coordinate?  What would the "rectangle" look like in that
case?  What would the formula say?  What about _both_ coordinates being the
same?
:::

### "Lifting" this to 3d

Now we have two points in the 3d space inside some rectangular coordinate
system, a point $P_1$ with coordinates $(x_1,y_1,z_1)$ and point
$P_2$ with coordinates $(x_2,y_2,z_2)$.  Instead of a rectangle, we will now
have a rectangular box with directly opposite corners at $P_1$ and $P_2$, and
with sides parallel to the three coordinate planes, and edges parallel to the
three coordinate axes.  _(Again, think about special cases when one, two or
even all three of the coordinates of the two points are equal.  How would the
"box" look like then?)_

```{r}
illustr("../../asymptote/distance/3ddistance.html", title = "Rectangular box with opposite corners at P1 and P2", width=285*2, height=256*2)
```

The distance between $P_1$ and $P_2$ is then the length of the diagonal
connecting the two points in the box.  To figure out this length, we can start
with a simpler question:

> What is the distance between the point $P_1$ with coordinates $(x_1, y_1,
> z_1)$ and the point $Q$ with coordinates $(x_2, y_2, z_1)$, that is the point
> that is the opposite corner to $P$ on the "bottom" side of the box?

Now both points are the opposite corners in the _same rectangle_, so we can use
the Pythagorean Theorem to find their distance.  We have a _right triangle_
where the hypotenuse is the diagonal $P_1Q$, and the sides have lengths
$\left\lvert x_2 - x_1\right\rvert$ and
$\left\lvert y_2 - y_1\right\rvert$.  From the Pythagorean Theorem we get the
length of the diagonal as
$$\sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2}$$
(Compare that with the 2d distance formula above! What is going on here?)

Now we have another triangle, with corners $P_1(x_1, y_1, z_1)$, $Q(x_2, y_2,
z_1)$ and $P_2(x_2, y_2, z_2)$.  This is also a right triangle (can you explain
why?).  You can rotate the illustration above to see that.  The hypotenuse of
this triangle is _the diagonal between $P_1$ and $P_2$_ whose length we are
trying to find! Applying the Pythagorean Theorem to this triangle will give us
the 3d distance formula.  Try it yourself and see if you can come up with the
following:

$$d = \sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2 + (z_2 - z_1)^2}$$

### Example


```{r}
vembedr::embed_youtube("Ld0iOipaKos")
```

### A bit of a stretch

If you compare the 2d and 3d distance formula, you will probably notice a
pattern.  In fact, this pattern extends even to the 1d case:

-   __In one dimension__, we calculate the distance between two points on the number
    line.  In your algebra classes, you have learned that the distance between
    two points $x_1$ and $x_2$ on the number line is the _absolute value of their
    difference_, or $\left\lvert x_2 - x_1\right\rvert$.

    you have probably also learned that the absolute value of a number can be
    written using square root of a square.  For any real number $a$, 

    $$\left\lvert a\right\rvert = \sqrt{a^2}$$

    We can then rewrite the 1d distance formula as

    $$\left\lvert x_2 - x_1\right\rvert = \sqrt{(x_2 - x_1)^2}$$

-   __In two dimensions__, the distance between two points $(x_1, y_1)$ and
    $(x_2, y_2)$ is

    $$\sqrt{(x_2 - x_1)^2 + (y_2 - y_2)^2}$$

-   __In three dimensions__, the distance between two points $(x_1, y_1, z_1)$ and
    $(x_2, y_2, z_2)$ is

    $$\sqrt{(x_2 - x_1)^2 + (y_2 - y_2)^2 + (z_2 - z_2)^2}$$


Can you make a conjecture how a distance between two points in a
_four-dimensional_ space would be calculated?  What about in 100 dimensions?

For a real stretch, what about infinitely dimensional space? (Think for example
about infinite sequences and series.  Things will get tricky, because you will
run into issues with convergence.  Not every infinite sequence will be useable
in that case.  This is way beyond the scope of this class, but you can look up
$\ell^2$ spaces, or even $L^2$ spaces if you are curious about this.)

### Try it yourself

```{r}
webwork("SVSU/math261/coordinates/distance_formula.pg", width=640, height=500)
```

