---
title: Equations in 3 (or more) variables
format: html
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

### Equations in 1 or 2 variables

From algebra, you know that a solution set of an equation in one variable is
a set of points on the number line $\mathbb{R}$. For example, the equation $x^2
- 3x + 2 = 0$ has solution set $\left\{1, 2\right\}$.

Similarly, a solution set of an equation in two variables is a set of points in
the plane $\mathbb{R}^2$. For example, the solutions set of the equation $x^2 +
y^2 = 1$ is the set of all points that form the circle with center $(0,0)$ and
radius 1.

```{r}
illustr("../../asymptote/equations/circle.html", title = "Circle with center (0,0) and radius 1", width=385, height=256)
```

Sometimes the same equation can be interpreted as an equation in one variable
or two variables: for example the equation $x = 2$ as an equation in one
variable has a single solution 2, while as an equation in two variables its set
of solutions consists of all the points in the form $(2,y)$ for any real $y$.
These points form a vertical line.

```{r}
illustr("../../asymptote/equations/line.html", title = "Vertical line through (2,0)", width=385, height=256)
```

### Equations in 3 variables

Each solution of an equation in three variables is a point in
the space $\mathbb{R}^3$. For example, one of the solutions of the equation $x
+ y + z = 2$ is the point $(1,1,0)$, another is $(\frac{1}{2}, \frac{1}{2},
1)$.

As another example, consider the equation $x^2 + y^2 + z^2 = 1$. We can rewrite
the equation as $$(x - 0)^2 + (y - 0)^2 + (z - 0)^2 = 1.$$

Taking the square root of both sides will produce $$\sqrt{(x-0)^2 + (y-0)^2 + (z-0)^2} = 1.$$

::: {.column-margin}
Why is it OK to take the square root of both sides? How do we know that the new equation is equivalent to the old one?
:::

The left hand side of the equation can be recognized as the _distance formula_
$$\sqrt{(x_2 - x_1)^2 + (y_2 - y_1)^2 + (z_2 - z_1)^2}$$
with $(x_2, y_2, z_2) = (x,y,z)$ and $(x_1, y_1, z_1) = (0,0,0)$. In other
words, the left hand side of the equation is __the distance of the point
$(x,y,z)$ from the origin__, and the equation says that this distance must be
equal to 1.

Therefore the solution set of the equation $x^2 + y^2 + z^2 = 1$ is the set of
all points in $\mathbb{R}^3$ that are exactly one unit away from the origin.
These points form the _sphere with center in the origin and radius 1_:

```{r}
illustr("../../asymptote/equations/sphere.html", title = "Sphere with center (0,0,0) and radius 1", width=385, height=256)
```

As another example, let's think about the equation $x = 2$ again, this time as
an equation in _three variables_. Any solution of this equation must look like
$(2, y, z)$, where $y$ and $z$ can be any real numbers (the equation does not
put any requirements on them at all).  The set of all such points form a
_plane_ that is parallel to the $yz$-plane (or, which is the same,
_perpendicular_ to the $x$-axis), passing through the point $(2,0,0)$:

```{r}
illustr("../../asymptote/equations/plane.html", title = "Vertical plane through (2,0,0)", width=385, height=256)
```

### Plotting equations using Sage

```{r}
#vembedr::embed_youtube("Ld0iOipaKos")
```

### Try it yourself

```{r}
webwork("Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch7/7-1-EquationOfASphere2.pg", width=640, height=500)
```

```{r}
webwork("Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch7/7-1-EquationOfASphere.pg", width=640, height=500)
```

```{r}
webwork("Library/Michigan/Chap12Sec1/Q11.pg", width=640, height=500)
```

```{r}
webwork("Library/Dartmouth/setStewartCh13S1/problem_3.pg", width=640, height=500)
```

