---
title: "Curl of a Vector Field"
format: html
layout: article
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    echo: false
    warning: false
---

{{< include ../_common.qmd >}}

## Curl of a vector field in $\mathbb{R}^3$

Previously we defined a _divergence_ of a vector field in $\mathbb{R}^2$ that
measures a local _expansion_ of the field. We then easily extended the formula
to $\mathbb{R}^3$.

Even earlier, we found a formula for calculating a local _circulation_ of a
vector field in $\mathbb{R}^2$. However, there does not seem to be any easy
obvious way how to extend that to $\mathbb{R}^3$. Here we will attempt to do
just that:

```{r echo=FALSE}
vembedr::embed_youtube("xmu6wCn4q6g")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap20Sec3/Q02.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/17_Fundamental_Theorems_of_Vector_Analysis/17.2_Stokes_Theorem/17.2.3.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap20Sec3/Q19.pg",
        width=640, height=500
)
```

