---
title: "Divergence of a Vector Field"
format: html
layout: article
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    echo: false
    warning: false
---

{{< include ../_common.qmd >}}

## Flux of a field across a curve

In this note it will be helpful to consider a vector field $\vec{F}(x,y) =
\v{P(x,y), Q(x,y)}$ as a _velocity field_ of a two-dimensional fluid flow.

Given such vector field $\vec{F}$ and a smooth curve $C$, we can ask _what is
the total volume_^[In two dimensions, _volume_ is the same as _area_.]
_of the fluid that crosses the curve from its left side to its
right side during one time unit_?

It will be easy to figure out in the case the curve is a _line segment_ and the
field is _constant_. The following diagram shows what happens then:

```{r,engine='tikz',fig.ext='svg',fig.width=5, engine.opts=list(dvisvgm.opts = "--font-format=woff")}
#| fig-cap: "Flow of constant field across a line segment in one time unit"
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{angles}
\usetikzlibrary{calc}
\usetikzlibrary{backgrounds}
\def\vec#1{\mathbf{#1}}
\begin{tikzpicture}
\begin{scope}[very thick,decoration={
    markings,
    mark=between positions 0.1 and 1 step 10mm with {\arrow{Stealth}}}
    ] 
    \draw[postaction={decorate}] (5,0) --node[below]{$C$} (0,3);
\end{scope}
\foreach \i in {0, 0.1, ..., 1.01}{
    \draw[thin, ->] (5,0) ++(-5*\i, 3*\i) -- +(2,1);
}
\coordinate (A) at (0,3);
\coordinate (B) at (5,0);
\coordinate (C) at ($(B)+(1.8,3)$);
\coordinate (D) at ($(B)+(2,1)$);
\coordinate (E) at ($(A)+(2,1)$);
\node[below right] at (D) {$\vec{F}$};
\draw[very thick, red, ->] (B) -- ($(B)!(D)!(C)$)node[right]{$\operatorname{proj}_{\vec{F}}\vec{N}$};
\draw[thick, dotted, ->] (B)pic[draw, solid, thin]{right angle = C--B--A} -- (C)node[above left]{$\vec{N}$};
\begin{scope}[on background layer]
    \fill[blue!30] (A) -- (B) -- (D) -- (E) -- cycle;
\end{scope}
\end{tikzpicture}
```

On the diagram^[Note that this diagram is _two-dimensional_: everything on this
picture is in the _same plane_!], the light blue area shows the amount of fluid that crosses the
line segment $C$ from left to right in one time unit. We can see that it is a
_parallelogram_, and its area can be calculated as the product of the length of
the curve $C$ and the "height" of the parallelogram. This height is equal to
the length of the _projection of $\vec{F}$ onto a unit normal vector $\vec{N}$
pointing to the right side of the curve, relative to the direction of travel
along the curve_.

In other words, the flux of the constant vector field $\vec{F}$ across the line
segment $C$ of length $L$ is equal to

$$L \comp_{\vec{N}} \vec{F} = L \frac{\vec{F}\cdot\vec{N}}{\norm{\vec{N}}} = L
\vec{F}\cdot \vec{N}$$

where $\vec{N}$ is the _unit normal vector_ to $C$ pointing to the _right side_
of $C$ with respect to the direction of travel along $C$.

### General field and curve

Suppose now $C$ is a _smooth_ curve and $\vec{F}$ is a _continuous_ vector
field defined at least on $C$. We can make the usual intuitive calculus
argument to derive the formula for the _flux of $\vec{F}$ across $C$_:

As always, we start by looking at a very tiny (possibly infinitesimal) piece of
$C$. Since $C$ is _smooth_, each tiny piece of $C$ is very close to a straight
line segment. Let's call the length of this tiny line segment $ds$. Since
$\vec{F}$ is _continuous_, we can consider it almost _constant_ when looking at
a very tiny piece of the curve. We can no apply the previously derived formula
to this tiny line segment and (almost) constant field $\vec{F}$: the flux of
$\vec{F}$ across this tiny line segment is approximately
$\vec{F}\cdot\vec{N}\;ds$.

We can do this for every "point" of the curve $C$, and add all the results
together, to get the following __flux integral__^[Here we are only barely
scratching the surface of this concept. Your textbook has more details about
flux integrals. We will come back to this concept later when talking about flux
integrals across surfaces in $\mathbb{R}^3$.]:

$$\int_C \vec{F}\cdot\vec{N}\;ds$$

where $\vec{N}$ is the _unit normal vector_ to $C$ pointing to the right side
of $C$ with respect to the direction of travel along $C$.

### Simple closed curves

Suppose now that $C$ is a positively oriented simple closed curve whose
interior region is $G$. Then $\vec{N}$ points __out__ away from $G$ towards the
exterior of $C$, so the flux of a field $\vec{F}$ across $C$ is equal to the
difference of the amount of fluid _leaving $G$_ and the amount of fluid
_entering $G$_. If this number is positive, it means that the fluid is somehow
_expanding_ or being created inside $G$, while if it is negative, the fluid is being
_compressed_ or annihilated inside $G$.

If for any simple closed curve the flux is 0, we say that the field $\vec{F}$
is _incompressible_.

This is kind of similar to the concept of _conservativeness_ except that
instead of _circulation integrals_ we use _flux integrals_. For circulation, we
were able to find a _local_ measure of circulation: $Q_x(x,y) - P_y(x,y)$, by
looking at a circulation along a tiny rectangle. We can try the same strategy
with flux.

Let's calculate a flux integral of a vector field $\vec{F} = \v{P, Q}$ over a very very tiny rectangle
centered at $(x_0,y_0)$, with sides of length $dx$ and $dy$. We will label the four
sides of the rectangle as $S_1$, $S_2$, $S_3$ and $S_4$:

```{r,engine='tikz',fig.ext='svg',fig.width=5, engine.opts=list(dvisvgm.opts = "--font-format=woff")}
#| fig-cap: "A tiny rectangle with normal vectors"
\usetikzlibrary{decorations.markings}
\def\vec#1{\mathbf{#1}}
\begin{tikzpicture}
\fill (1.5, 1.5) circle[radius=1pt] node[below]{$(x_0,y_0)$};
\begin{scope}[thick,decoration={
    markings,
    mark=at position 0.7 with {\arrow{>}}}
    ] 
    \draw[postaction={decorate}] (0,0) --node[above]{$S_1$} (3,0);
    \draw[postaction={decorate}] (3,0) --node[left]{$S_2$} (3,3);
    \draw[postaction={decorate}] (3,3) --node[below]{$S_3$} (0,3);
    \draw[postaction={decorate}] (0,3) --node[right]{$S_4$} (0,0);
\end{scope}
\fill (0,0) circle[radius=1pt] node[below left]{$(x_0-\frac{dx}{2},y_0-\frac{dy}{2})$};
\fill (3,0) circle[radius=1pt] node[below right]{$(x_0+\frac{dx}{2},y_0-\frac{dy}{2})$};
\fill (3,3) circle[radius=1pt] node[above right]{$(x_0+\frac{dx}{2},y_0+\frac{dy}{2})$};
\fill (0,3) circle[radius=1pt] node[above left]{$(x_0-\frac{dx}{2},y_0+\frac{dy}{2})$};
\draw[thick, red, ->] (3,1.5) -- (4,1.5)node[right]{$\vec{N}_2 = \left\langle 1,0\right\rangle$};
\draw[thick, red, ->] (0,1.5) -- (-1,1.5)node[left]{$\vec{N}_4 = \left\langle -1,0\right\rangle$};
\draw[thick, red, ->] (1.5,3) -- (1.5,4)node[above]{$\vec{N}_3 = \left\langle 0,1\right\rangle$};
\draw[thick, red, ->] (1.5,0) -- (1.5,-1)node[below]{$\vec{N}_1 = \left\langle 0,-1\right\rangle$};
\end{tikzpicture}
```

Then the flux integral of $\vec{F}$ across the rectangle can be split into the sum of
flux integrals over the four sides:

$$\int_{S_1} \vec{F}(x,y)\cdot \vec{N}_1\;ds + \int_{S_2} \vec{F}(x,y)\cdot \vec{N}_2\;ds + \int_{S_3} \vec{F}(x,y)\cdot \vec{N}_3\;ds + \int_{S_4} \vec{F}(x,y)\cdot \vec{N}_4\;ds $$

Plugging in the four normal vectors $\vec{N}_1 = \v{0,-1}$, $\vec{N}_2 =
\v{1,0}$, $\vec{N}_3 = \v{0,1}$ and $\vec{N}_4 = \v{-1,0}$ and calculating the
dot products will give us

$$\int_{S_1} -Q(x,y)\;ds + \int_{S_2} P(x,y)\;ds + \int_{S_3} Q(x,y)\;ds + \int_{S_4} -P(x,y)\;ds$$

Pulling the minus signs out of the integrals and changing the order will give
us

$$\int_{S_3} Q(x,y)\;ds - \int_{S_1} Q(x,y)\;ds + \int_{S_2} P(x,y)\;ds - \int_{S_4} P(x,y)\;ds$$

When the rectangle is very small, we can approximate each of the integrals by
the value in the middle of the segment times the length of the segment:

$$Q\left(x_0, y_0 + \frac{dy}{2}\right)\;dx - Q\left(x_0, y_0 - \frac{dy}{2}\right)\;dx + P\left(x_0 + \frac{dx}{2}, y_0\right)\;dy - P\left(x_0 - \frac{dx}{2}, y_0\right)\;dy$$

Then $Q\left(x_0, y_0 + \frac{dy}{2}\right) - Q\left(x_0, y_0 - \frac{dy}{2}\right)$
can be approximated by $Q_y(x_0, y_0)\;dy$ and 
$P\left(x_0 + \frac{dx}{2}, y_0\right) - P\left(x_0 - \frac{dx}{2}, y_0\right)$
can be approximated by $P_x(x_0, y_0)\;dx$.

Altogether, the flux across the tiny rectangle is approximately equal to

$$\left(P_x(x_0, y_0) + Q_y(x_0, y_0)\right)\;dx\;dy$$

This is very similar to the expression for local circulation along the tiny
rectangle, which was 

$$\left(Q_x(x_0, y_0) - P_y(x_0, y_0)\right)\;dx\;dy$$

Notice that for the circulation, we had the derivative of the __second__
component with respect to the __first__ variable __minus__ the derivative of
the __first__ component with respect to the __second__ variable.

For flux, we have the derivative of the __first__ component with respect to the
__first__ variable __plus__ the derivative of the __second__ component with
respect to the __second__ variable.

## Divergence

Given a vector field $\vec{F}(x,y) = \v{P(x,y), Q(x,y)}$, the (scalar) function 
$$P_x(x, y) + Q_y(x, y)$$
is called the __divergence__ function of the field $\vec{F}$, and is denoted as
$\divg{F}$.

This can be naturally extended to three (or, in fact, any number of)
dimensions: given 
$$\vec{F}(x,y,z) = \v{P(x,y,z), Q(x,y,z), R(x,y,z)}$$
the divergence of $\vec{F}$ is defined as
$$\divg{F}(x,y,z) = P_x(x,y,z) + Q_y(x,y,z) + R_z(x,y,z).$$

### Examples

```{r echo=FALSE}
vembedr::embed_youtube("4PWZ6xL3E0E")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap20Sec1/Q07.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/17_Fundamental_Theorems_of_Vector_Analysis/17.3_Divergence_Theorem/17.3.3.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/20-1-Idea-of-divergence/HGM4-20-1-09-Divergence.pg",
        width=640, height=500
)
```
