---
title: "Operator Notation, Divergence and Curl"
format: html
layout: article
reference-location: margin
cap-location: margin
knitr:
  opts_chunk: 
    echo: false
    warning: false
---

{{< include ../_common.qmd >}}

## Operator notation

In the context of calculus, an _operator_ is a function that takes functions as
inputs and produce either a function or a number as an output.

Some examples:

*   _The identity operator_ $I$: it receives a function and produces the exact
    same function as output.
*   _First derivative of a real function of one variable_: takes a differentiable
    function and produces a function.
*   _Integral over a given interval_: takes an integrable function and produces
    a number.
*   _An indefinite integral with a fixed lower limit_: takes an integrable
    function and produces a function.
*   _Partial derivative of a function of two variables with respect to $x$_:
    takes a function of two variables and produces a function of two variables.
*   _Gradient of a function of two variables_: takes a function of two
    variables and produces a two-dimensional vector field.
*   _Curvature_: takes a vector function of single variable and produces a real
    function of single variable.

We usually denote operators using uppercase letters, Greek letters, or special
symbols. Applying an operator to a function is usually denoted _without
parentheses_, unless order of operations must be specified, and is considered a _form of multiplication_.

For example, a simple first derivative operator would be defined as

$$D f = f'$$

Then $D \sin$ would be $\cos$ and so on.

Operators can generally be _added_ and _multiplied by a scalar_. Given two
operators, $A$ and $B$, the operator $2A + 3B$ would be defined the following
way:

$$(2A + 3B)f = 2(Af) + 3(Bf)$$

where the addition and scalar multiplication on the right side of the equality
is a regular addition and scalar multiplication of functions.

In contrast to that, _multiplication_ or operators usually means
__composition__, so $(AB)f = A(Bf)$. For example, taking the first derivative
operator $D$ defined above and multiplying it by itself would give us a
derivative of derivative, which is the _second derivative_. Note that while
composition in general is _not commutative_, when composing an operator with
itself, the order clearly does not matter, so the multiplication of $D$ by $D$
can be abbreviated as $D^2$. 

For example, the operator $D^2 + 2D + 5I$ would take a second derivative of a
function, add to it 2 times the first derivative of the same function, and add
5 times the same function to that.  In other words,
$$(D^2 + 2D + 5I)f = f'' + 2f' + 5f$$

### Partial derivative operators

When working with functions from $\mathbb{R}^2$ to $\mathbb{R}$, we have two
_partial derivative operators_:

*   Partial derivative with respect to $x$, denoted either as
    $\frac{\partial}{\partial x}$ or $D_x$.
*   Partial derivative with respect to $y$, denoted either as
    $\frac{\partial}{\partial y}$ or $D_y$.

In $\mathbb{R}^3$, we also have the partial derivative with respect to $z$, denoted either as
$\frac{\partial}{\partial z}$ or $D_z$.

This can of course be extended to any number of variables.

### Vectors of operators

We can form a symbolic vector whose components are operators. Applying such
operator to a function would work similarly to scalar multiplication, except
that contrary to the usual convention, the "scalar" would be on the right side
of the vector instead of the left side.  The result of such operator would then
be a _vector function_.  For example, an operator $\vec{V} = \v{D, D^2,
D^3}$, when applied to a function $f$, would produce a vector function whose first
component would be $f'$, second component $f''$ and third component $f'''$.

### Gradient operator

An important example of an operator vector is the _gradient operator_. In
$\mathbb{R}^2$, it is defined as
$$\vec{\nabla} = \v{\frac{\partial}{\partial x}, \frac{\partial}{\partial y}}$$
In $\mathbb{R}^3$, it is
$$\vec{\nabla} = \v{\frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z}}$$

### Example: the Laplace operator

A very important operator in mathematical physics is the _Laplace operator_
$\Delta$,
also known as _Laplacian_. It can be defined in a number of different ways, one
of them being the _dot product of the gradient operator with itself_!

$$\Delta = \vec{\nabla}\cdot\vec{\nabla}$$

In fact, the Laplace operator is sometimes also denoted as $\nabla^2$.

How does it work? Since $\vec{\nabla} = \v{\frac{\partial}{\partial x},
\frac{\partial}{\partial y}}$, and since _dot product_ works by multiplying the
corresponding components and adding all the products together,

\begin{align}
\Delta &= \vec{\nabla}\cdot\vec{\nabla}\\
&= \v{\frac{\partial}{\partial x}, \frac{\partial}{\partial y}}\cdot\v{\frac{\partial}{\partial x}, \frac{\partial}{\partial y}}\\
&= \frac{\partial}{\partial x}\frac{\partial}{\partial x} + \frac{\partial}{\partial y}\frac{\partial}{\partial y}\\
&= \frac{\partial^2}{(\partial x)^2} + \frac{\partial^2}{(\partial y)^2}
\end{align}

So the Laplace operator _takes the second pure partial derivatives and adds them
together_:

$$\Delta f(x,y) = f_{xx}(x,y) + f_{yy}(x,y)$$

The same can be done in $\mathbb{R}^3$ (or in any number of dimensions):
$$\Delta f(x,y,z) = f_{xx}(x,y,z) + f_{yy}(x,y,z) + f_{zz}(x,y,z)$$

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart16_5/problem_3.pg",
        width=640, height=500
)
```

## Divergence and Curl Operators

```{r echo=FALSE}
vembedr::embed_youtube("CtkIHuyiAKs")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap20Sec5/Q01.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Dartmouth/setMTWCh4S4/problem_5.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Rochester/setVectorCalculus2/ur_vc_12_10.pg",
        width=640, height=500
)
```

