---
title: Cross Product
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## Determinants

Before learning about cross product, it is useful to review some basic
information about _determinants_:

```{r echo=FALSE}
vembedr::embed_youtube("reNSHrJclCA")
```

## In two dimensions

(Note: _According to some texts, the operation of __cross product__ is not
actually defined in two dimensions, it is only defined in three dimensions.
The 2d vector operation described here is often called a __perp dot product__
for reasons described below._)

The __cross product__ of two 2-dimensional vectors $\vec{u} = \v{u_1, u_2}$ and
$\vec{v} = \v{v_1, v_2}$ is defined as the scalar

$$\vec{u}\times\vec{v} = \norm{\vec{u}}\norm{\vec{v}}\sin\theta$$

where $\theta$ is the angle between $\vec{u}$ and $\vec{v}$.

It is equal to the "signed area" of the parallelogram defined by the two
vectors:

```{r, echo=FALSE}
illustr("../../asymptote/vectors/2dcross.html", title = "Cross product as an area", width=285, height=256)
```

(The area of a parallelogram is equal to the base times height.  In the
parallelogram above, the base is $\norm{\vec{u}}$ and the height is
$\norm{\vec{v}}\sin\theta$.  If $\theta <0$, we get a "negative" area".)

In components, it turns out the cross product or vectors $\vec{u} = \v{u_1,
u_2}$ and $\vec{v} = \v{v_1, v_2}$ is

$$\vec{u}\times\vec{v} = u_1v_2 - v_1u_2$$

### Explanation and connection to dot product:

The following illustration shows two vectors: $\vec{u}$ and $\vec{v}$, together
with a vector $\vec{u}^\perp$, which is a vector of the same magnitude as
$\vec{u}$, but _perpendicular_ to $\vec{u}$.  If the component form of
$\vec{u}$ is $\v{u_1,u_2}$, the comoponent form of $\vec{u}^\perp$ is $\v{-u_2,
u_1}$ (You can also use $\v{u_2, -u_1}$, which would point in the opposite
direction.  You can verify that you will get the same result in the end.)

```{r, echo=FALSE}
illustr("../../asymptote/vectors/perpdot.html", title = "Vector perpendicular to u", width=285, height=256)
```

Then 

$$\begin{aligned}
\vec{u}^\perp \cdot \vec{v} &= \norm{\vec{u}^\perp}\norm{\vec{v}}\cos\left(\frac{\pi}{2}-\theta\right)\\
&=\norm{\vec{u}}\norm{\vec{v}}\sin\left(\theta\right)\\
&=\vec{u}\times\vec{v}
\end{aligned}$$

In components, $\vec{u}^\perp \cdot \vec{v} = -u_2v_1 + u_1v_2 = u_1v_2
- u_2v_1$.

This is why the "2d cross product" is sometimes called a _perp dot product_.
The notation is then $\vec{u}\perp\vec{v}$.  Another term used for this
operation is the _outer product_.

### Properties

From the definition and the equivalent component formula, we get the following
properties:

-   __Anti-commutativity:__ $\vec{u}\times\vec{v} = -\vec{v}\times\vec{u}$

-   __Nilpotent property:__ $\vec{u}\times\vec{u} = 0$

-   __Distributive property:__ $\vec{u}\times\left(\vec{v} + \vec{w}\right) =
  (\vec{u}\times\vec{v}) + (\vec{u}\times\vec{w})$

-   __Associativity with scalar multiple:__ $c(\vec{u}\times\vec{v} =
  (c\vec{u})\times\vec{v} = \vec{u}\times(c\vec{v})$

-  __Lagrange property:__ $(\vec{u}\times\vec{v})^2 + (\vec{u}\cdot\vec{v})^2 =
  (\norm{u}\norm{v})^2$

(The last one follows directly from the Pythagorean identity for sine and
cosine.)

### Calculating using _determinants_

The 2d cross product $\vec{u}\times\vec{v}$ can also be defined as a determinant of a matrix with 
first row $\vec{u}$ and second row $\vec{v}$:

$$\vec{u}\times\vec{v} = \left\lvert \begin{matrix} u_1 & u_2\\ v_1 & v_2 \end{matrix} \right\rvert = u_1v_2 - u_2v_1$$


### Example

> Find the area of the parallelogram defined by the vectors $\vec{a} = \v{3,2}$
> and $\vec{b} = \v{1,-5}$.

The _signed_ area of the parallelogram is

$$\vec{a}\times \vec{b} = \left\lvert\begin{matrix}3 & 2\\1 &
-5\end{matrix}\right\rvert = 3\cdot (-5) - 2\cdot 1 = -17$$

so the area is 17 square units.

### Cross product in 3 dimensions

The true cross product of two vectors is defined when the two vectors are 3d
vectors.  The _cross product_ of two 3d vectors $\vec{u}$ and $\vec{v}$ (also known as the _vector
product_ or _outer product_) is a __vector__ with the following properties:

-  It is _perpendicular_ to both $\vec{u}$ and $\vec{v}$.
-  Its magnitude is equal to the area of the parallelogram defined by the
  vectors $\vec{u}$ and $\vec{v}$.
-  The direction is determined by the _right hand rule_.

```{r echo=FALSE}
illustr("../../asymptote/vectors/cross3d.html", title = "Cross product in 3d",
        width = 2*285, height = 2*256)
```

In components, the cross product can be calculated using the following
determinant:

\begin{aligned}
\vec{u}\times\vec{v} =
\left\lvert\begin{matrix}
\vec{i} & \vec{j} & \vec{k}\\
u_1 & u_2 & u_3\\
v_1 & v_2 & v_3
\end{matrix}\right\rvert &= \v{\left\lvert\begin{matrix}u_2 & u_3\\v_2 &
v_3\end{matrix}\right\rvert, -\left\lvert\begin{matrix}u_1 & u_3\\v_1 &
v_3\end{matrix}\right\rvert, \left\lvert\begin{matrix}u_1 & u_2\\v_1 &
v_2\end{matrix}\right\rvert}\\
&= \v{u_2v_3 - u_3v_2, -(u_1v_3 - u_3v_1), u_1v_2 - u_2v_1}
\end{aligned}

#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart12_4/problem_1.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Rochester/setVectors3CrossProduct/ur_vc_2_3.pg",
        width=640, height=500
)
```


### Properties

-   __Anti-commutativity:__ $\vec{u}\times\vec{v} = -\vec{v}\times\vec{u}$

-   __Nilpotent property:__ $\vec{u}\times\vec{u} = \v{0,0,0} = \vec{O}$

-   __Zero vector product:__ $\vec{O}\times\vec{u} = \vec{u}\times\vec{O} = \vec{O}$

-   __Distributive property:__ $\vec{u}\times\left(\vec{v} + \vec{w}\right) =
  (\vec{u}\times\vec{v}) + (\vec{u}\times\vec{w})$

-   __Associativity with scalar multiple:__ $c(\vec{u}\times\vec{v} =
  (c\vec{u})\times\vec{v} = \vec{u}\times(c\vec{v})$

-  __Lagrange property:__ $\norm{\vec{u}\times\vec{v}}^2 + (\vec{u}\cdot\vec{v})^2 =
  (\norm{u}\norm{v})^t2$

-  __Orthogonality:__ $(\vec{u}\times\vec{v})\cdot\vec{u} = (\vec{u}\times\vec{v})\cdot\vec{v} = 0$

-  __Triple scalar product:__ $\vec{u}\cdot(\vec{v}\times\vec{w}) = (\vec{u}\times\vec{v})\cdot\vec{w}$

The first three properties follow directly from the properties of determinants:

-   Switching two rows in a determinant changes the sign of the determinant

-   Determinant with two equal rows is 0

-   Determinant with a row of zeros is 0

The Lagrange property is a consequence of the Pythagorean identity for sine and
cosine.  The rest of the properties can be proven by working with components.

For us, the most useful properties will be the __orthogonality__ and the fact
that $\norm{\vec{u}\times\vec{v}}$ is __the area of the parallelogram__ defined
by $\vec{u}$ and $\vec{v}$.

#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/UMN/calculusStewartET/s_12_4_24.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart12_4/problem_3.pg",
        width=640, height=500
)
```

### Triple Scalar Product

The triple scalar product $\vec{u}\cdot(\vec{v}\times\vec{w})$ is related to
volume on the following way:

```{r echo=FALSE}
vembedr::embed_youtube("FEPw8PzUKmE")
```

It also gives another interpretation to a $3\times 3$ determinant.  The
determinant

$$\left\lvert\begin{matrix} a_1 & a_2 & a_3\\ b_1 & b_2 & b_3\\ c_1 & c_2 & c_3\\ \end{matrix}\right\rvert$$

is the _signed volume_ of the _parallelepiped_ determined by the vectors 
$\vec{a} = \v{a_1, a_2, a_3}$, 
$\vec{b} = \v{b_1, b_2, b_3}$, and
$\vec{c} = \v{c_1, c_2, c_3}$,
where the sign is determined by the order in which the vectors are listed.


#### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/UMN/calculusStewartET/s_12_4_prob01.pg",
        width=640, height=500
)
```

