---
title: Dot Product
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

## Two ways to define

### Vectors as "Arrows"

The _angle between vectors $\vec{u}$ and $\vec{v}$_ in $\mathbb{R}^2$ or
$\mathbb{R}^3$ is simply the angle,
between $-\pi$ and $\pi$, one would need to rotate the vector $\vec{u}$ by so that
it would become parallel to the vector $\vec{v}$:

```{r, echo=FALSE}
illustr("../../asymptote/vectors/angle.html", title = "Angle between two vectors", width=285, height=256)
```

On the diagram above, $\theta$ is the angle between $\vec{u}$ and $\vec{v}$.

The __dot product__ (also called _scalar product_ or _inner product_) of two vectors $\vec{u}$ and
$\vec{v}$ in $\mathbb{R}^2$ or $\mathbb{R}^3$ is the scalar quantity

$$\vec{u}\cdot\vec{v} = \norm{\vec{u}}\norm{\vec{v}}\cos\theta$$

where $\theta$ is the angle between $\vec{u}$ and $\vec{v}$.

#### Some properties

-   The dot product is _commutative_: $\vec{u}\cdot\vec{v} =
    \vec{v}\cdot\vec{u}$.  This follows from the fact that $\cos$ is an _even_
    function, so $\cos(-\theta) = \cos(\theta)$.

-   The dot product $\vec{u}\cdot\vec{v}$ is _zero_ if and only if at least one
    of the following is true:

    -   $\vec{u}$ is the zero vector
    -   $\vec{v}$ is the zero vector
    -   $\vec{u}$ and $\vec{v}$ are _perpendicular_ to each other.

    This means that the dot product does not satisfy the _zero product
    property_: dot product of two vectors can be zero even when none of the two
    vectors are zero!

-   The dot product $\vec{u}\cdot\vec{v}$ is _positive_ if and only if the angle
    between the two vectors is _acute_, that is between $-\frac{\pi}{2}$ and
    $\frac{\pi}{2}$.

-   The dot product $\vec{u}\cdot\vec{v}$ is _negative_ if and only if the angle
    between the two vectors is _obtuse_.

*   For any scalar^[This is true even for a _negative_ $c$! Think about what happens to an
    angle between two vectors when one of the vectors is multiplied by a
    negative scalar!] $c$, $(c\vec{u})\cdot\vec{v} = \vec{u}\cdot(c\vec{v}) =
    c(\vec{u}\cdot\vec{v})$

-   $\vec{u}\cdot\vec{v} = \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 - \norm{\vec{u} - \vec{v}}^2\right)$

    This is a consequence of the _law of cosines_.  Applying the law of cosines
    on the following triangle:

    ```{r}
    illustr("../../asymptote/vectors/cosines.html", title = "Triangle defined by vectors u and v", width=285, height=256)
    ```

    will give us $\norm{\vec{u} - \vec{v}}^2 = \norm{\vec{u}}^2 +
    \norm{\vec{v}}^2 - 2\norm{\vec{u}}\norm{\vec{v}}\cos\theta$.

-   $\vec{u}\cdot\vec{u} = \norm{\vec{u}}^2$.


#### Try it yourself

```{r}
webwork("Library/Rochester/setVectors2DotProduct/UR_VC_1_11.pg", width=640, height=500)
```

### Vectors in components

If $\vec{u}$ and $\vec{v}$ are two vectors in $\mathbb{R}^2$ or $\mathbb{R}^3$,
we know that (as a consequence of the law of cosines)

$$\vec{u}\cdot\vec{v} = \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 - \norm{\vec{u} - \vec{v}}^2\right).$$

Writing $\vec{u} = \v{u_1, u_2, u_3}$ and $\vec{v} = \v{v_1, v_2, v_3}$ in
$\mathbb{R}^3$ (or similarly with just two components in $\mathbb{R}^2$), the
identity becomes
$$\begin{aligned}
\vec{u}\cdot\vec{v} &= \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 -
\norm{\vec{u} - \vec{v}}^2\right)\\
&= \frac{1}{2}\left(u_1^2 + u_2^2 + u_3^2 + v_1^2 + v_2^2 + v_3^2 - (u_1 - v_1)^2 - (u_2 - v_2)^2 - (u_3 - v_3)^2\right)\\
\end{aligned}$$
Expanding the squares and combining like terms will turn this into
$$\frac{1}{2}\left(2u_1v_1 + 2u_2v_2 + 3u_3v_3\right) = u_1v_1 + u_2v_2 + u_3v_3$$
which gives us an easy way to calculate a dot product of $\vec{u}$ and
$\vec{v}$ using their components.

#### Dot product in any finite dimension

In light of this, we can now define dot product between two vectors $\vec{u}$
and $\vec{v}$ in $\mathbb{R}^n$ for any integer $n > 0$:

Given
$$\vec{u} = \v{u_1, u_2, \dots, u_n} \in \mathbb{R}^n$$
and
$$\vec{v} = \v{v_1, v_2, \dots, v_n} \in \mathbb{R}^n$$
define
$$\vec{u}\cdot\vec{v} = u_1v_1 + u_2v_2 + \cdots + u_nv_n = \sum_{i=1}^n u_iv_i.$$

:::{.column-margin}
Again, just as with the norm, this can be extended to infinitely dimensional
vectors, such as sequences or functions, as long as one takes care of
convergence issues in some way.
:::

*   If $n = 1$, this just becomes a regular product of two numbers.
*   As we have just shown, if $n = 2$ or $n = 3$, this agrees with the earlier
    definition of dot product using an angle between vectors.
*   In higher dimensions, it gives us a way to define a dot product even if
    there is no intuitive geometrical concept of an angle between vectors.

##### Example:

If $\vec{u} = \v{1,-2,4,3}$ and $\vec{v} = \v{2,5,-1,2}$ then
$$\vec{u}\cdot\vec{v} = 1\cdot 2 + (-2)\cdot 5 + 4\cdot (-1) + 3\cdot 2 = 2 - 10 - 4 + 6 = -6.$$

#### Try it yourself

```{r}
webwork("Library/Union/setMVvectors/vectors-6.pg", width=640, height=500)
```

### Properties of dot product

Earlier we saw several properties of dot product in $\mathbb{R}^2$ and
$\mathbb{R}^3$ that followed easily from the
geometric definition. Most of these properties are still true in any
dimension, and can be proved by rewriting vectors using components and
manipulating the resulting expressions using rules of algebra.  In addition to
that, there are also some other properties that are very hard to prove from the
geometric definition, but became relatively easy when using components. Here is
a list with some important properties of dot product:

-   The dot product is _commutative_: $\vec{u}\cdot\vec{v} =
    \vec{v}\cdot\vec{u}$.

-   Dot product does __not__ satisfy the _zero product
    property_: dot product of two vectors can be zero even when none of the two
    vectors are zero!

-   Dot product _distributes_^[The distributive property is hard to prove geometrically, but very easy to prove using
    components! Give it a try!] over vector addition: $\vec{u}\cdot(\vec{v} +
    \vec{w}) = \vec{u}\cdot\vec{v} + \vec{u}\cdot\vec{w}$.


*   For any scalar $c$, $(c\vec{u})\cdot\vec{v} = \vec{u}\cdot(c\vec{v}) =
    c(\vec{u}\cdot\vec{v})$

-   $\vec{u}\cdot\vec{v} = \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 - \norm{\vec{u} - \vec{v}}^2\right)$

-   $\vec{u}\cdot\vec{u} = \norm{\vec{u}}^2$.

-   __Cauchy--Schwartz inequality__: ^[A nice simple proof of this version of Cauchy--Schwartz inequality can be found
    for example on
    [Wikipedia](https://en.wikipedia.org/w/index.php?title=Cauchy%E2%80%93Schwarz_inequality&oldid=1085897952#Rn_-_n-dimensional_Euclidean_space).] for any two vectors $\vec{u}$ and
    $\vec{v}$ in $\mathbb{R}^n$, $\abs{\vec{u}\cdot\vec{v}} \le
    \norm{\vec{u}}\norm{\vec{v}}$.

### Angles between vectors

In $\mathbb{R}^2$ and $\mathbb{R}^3$, we have two equivalent ways to calculate
the dot product of two vectors: one using components and one using the angle
between the vectors. If we know the components of the two vectors, we can use
this to calculate the angle, or at least its absolute value!

Starting with the equation $\vec{u}\cdot\vec{v} =
\norm{\vec{u}}\norm{\vec{v}}\cos\theta$ and solving for $\cos\theta$, we get

$$\cos\theta = \frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}$$

Using the inverse cosine we get

$$\abs{\theta} = \cos^{-1}\left(\frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}\right)$$

for any angle $\theta$ between $-\pi$ and $\pi$.

#### Example

Let's take two vectors, $\vec{u} = \v{1,3,2}$ and $\vec{v} = \v{2,-2,1}$.
We get

$$\begin{aligned}
\norm{\vec{u}} &= \sqrt{1^2 + 3^2 + 2^2} = \sqrt{14}\\
\norm{\vec{v}} &= \sqrt{2^2 + (-2)^2 + 1^2} = \sqrt{9} = 3\\
\vec{u}\cdot\vec{v} &= 1\cdot 2 + 3\cdot(-2) + 2\cdot 1 = 1 - 6 + 1 = -4
\end{aligned}$$

so

$$\cos\theta = \frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}} = \frac{-4}{3\sqrt{14}}$$

That means that either

$$\theta = \cos^{-1}\left(-\frac{4}{3\sqrt{14}}\right)$$

or

$$\theta = -\cos^{-1}\left(-\frac{4}{3\sqrt{14}}\right)$$

Either way, $\theta$ is an _obtuse_ angle.

:::{.column-margin}
Note that while in $\mathbb{R}^2$, the sign of an angle is always well defined:
counter-clockwise rotations are positive and clockwise are negative, in
$\mathbb{R}^3$ the sign of an angle depends on which side we are viewing the
angle from.
:::

### Angles in higher dimensions

In fact, we can use dot product to define both angles and distances in higher
dimensions.

First, we can define the norm of a vector as $\norm{\vec{u}} =
\sqrt{\vec{u}\cdot\vec{u}}$. Once we know how to find a norm of a vector, we
can define the distance between two points $P$ and $Q$ as the norm of
the vector from $P$ to $Q$.

Second, we can define the angle $\theta$ between two non-zero vectors $\vec{u}$ and
$\vec{v}$ as
$$\theta = \cos^{-1}\left(\frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}\right)$$

This is well defined: according to the Cauchy--Schwartz inequality,
$$-1 \le \frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}} \le 1$$
which means that 
$$\cos^{-1}\left(\frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}\right)$$
is defined and between $0$ and $\pi$.

### Orthogonal vectors

We saw that there are two ways for $\vec{u}\cdot\vec{v}$ to be 0:

1.   One or both of the vectors are _zero vectors_.
2.   Both vectors are non-zero and the angle between the vectors is
    $\frac{\pi}{2}$ or $-\frac{\pi}{2}$, in other words, the vectors are
    _perpendicular_ to each other.

To make things easier, we define the concept of __orthogonal__ vectors:

We say that vectors $\vec{u}$ and $\vec{v}$ are _orthogonal to each other_, or
that $\vec{u}$ is _orthogonal_ to $\vec{v}$, or that $\vec{v}$ is _orthogonal_
to $\vec{u}$, if the dot product $\vec{u}\cdot\vec{v} = 0$.

For non-zero vectors, _orthogonal_ is the same as _perpendicular_.  The zero
vector is orthogonal to all vectors, including itself!

### Try it yourself

```{r}
webwork("Library/Union/setMVvectors/vectors-7.pg", width=640, height=500)
```

```{r}
webwork("Library/272/setStewart12_3/problem_1a.pg", width=640, height=500)
```

```{r}
webwork("Library/Michigan/Chap13Sec3/Q31.pg", width=640, height=500)
```

```{r}
webwork("Library/UMN/calculusStewartET/s_12_3_22.pg", width=640, height=500)
```
