---
title: Dot Product
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

## Two ways to define

### Vectors as "Arrows"

The _angle between vectors $\vec{u}$ and $\vec{v}$_ is simply the angle,
between $-\pi$ and $\pi$, one would need to rotate the vector $\vec{u}$ so that
it would become parallel to the vector $\vec{v}$:

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/angle.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Angle between two vectors", src = basename(fname), width=285, height=256, frameborder="0")
```

On the diagram above, $\theta$ is the angle between $\vec{u}$ and $\vec{v}$.

The __dot product__ (also called _scalar product_ or _inner product_) of two vectors $\vec{u}$ and
$\vec{v}$ is the scalar quantity

\[\vec{u}\cdot\vec{v} = \norm{\vec{u}}\norm{\vec{v}}\cos\theta\]

where $\theta$ is the angle between $\vec{u}$ and $\vec{v}$.

#### Some properties

-   The dot product is _commutative_: $\vec{u}\cdot\vec{v} =
  \vec{v}\cdot\vec{u}$.  This follows from the fact that $\cos$ is an _even_
  function, so $\cos(-\theta) = \cos(\theta)$.

-  The dot product $\vec{u}\cdot\vec{v}$ is _zero_ if and only if at least one
  of the following is true:

    -   $\vec{u}$ is the zero vector
    -   $\vec{v}$ is the zero vector
    -   $\vec{u}$ and $\vec{v}$ are _perpendicular_ to each other.

-  The dot product $\vec{u}\cdot\vec{v}$ is _positive_ if and only if the angle
  between the two vectors is _acute_, that is between $-\frac{\pi}{2}$ and
  $\frac{\pi}{2}$.

-  The dot product $\vec{u}\cdot\vec{v}$ is _negative_ if and only if the angle
  between the two vectors is _obtuse_.

-  $\vec{u}\cdot\vec{v} = \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 - \norm{\vec{u} - \vec{v}}^2\right)$

    This is a consequence of the _law of cosines_.  Applying the law of cosines
    on the following triangle:

    ```{r, echo=FALSE}
    fname <- "../../asymptote/vectors/cosines.html"
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = "Triangle defined by vectors u and v", src = basename(fname), width=285, height=256, frameborder="0")
    ```

    will give us $\norm{\vec{u} - \vec{v}}^2 = \norm{\vec{u}}^2 +
    \norm{\vec{v}}^2 - 2\norm{\vec{u}}\norm{\vec{v}}\cos\theta$.

-   $\vec{u}\cdot\vec{u} = \norm{\vec{u}}^2$.


#### Try it yourself

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/Rochester/setVectors2DotProduct/UR_VC_1_11.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


### Vectors in coordinates

Let's introduce a coordinate system, so that $\vec{u} = \v{u_1, u_2, u_3}$ and
$\vec{v} = \v{v_1, v_2, v_3}$.  We already know that 

-  $\norm{\vec{u}} = \sqrt{u_1^2 + u_2^2 + u_3^2}$
-  $\norm{\vec{v}} = \sqrt{v_1^2 + v_2^2 + v_3^2}$
-  $\norm{\vec{u - v}} = \sqrt{(u_1 - v_1)^2 + (u_2 - v_2)^2 + (u_3 - v_3)^2}$

Then 

\[\begin{aligned}
   \vec{u}\cdot\vec{v} &= \frac{1}{2}\left(\norm{\vec{u}}^2 + \norm{\vec{v}}^2 - \norm{\vec{u} - \vec{v}}^2\right)\\
   &= \frac{1}{2}\left(u_1^2 + u_2^2 + u_3^2 + v_1^2 + v_2^2 + v_3^2 - (u_1 - v_1)^2 - (u_2 - v_2)^2 - (u_3 - v_3)^2\right)\\
   &= \frac{1}{2}\left(u_1^2 + u_2^2 + u_3^2 + v_1^2 + v_2^2 + v_3^2 - u_1^2 + 2u_1v_1 - v_1^2 - u_2^2 + 2u_2v_2 - v_2^2 - u_3^2 + 2u_3v_3 - v_3^2\right)\\
   &= \frac{1}{2}\left(2u_1v_1 + 2u_2v_2 + 2u_3v_3\right)\\
   &= u_1v_1 + u_2v_2 + u_3v_3
\end{aligned}\]

#### Definition

Given two vectors, $\vec{u} = \v{u_1, u_2, u_3}$ and $\vec{v} = \v{v_1, v_2, v_3}$, define
\[\vec{u}\cdot\vec{v} = u_1v_1 + u_2v_2 + u_3v_3\]

From the above it follows that both definitions are equivalent.

#### Some additional properties

Some properties that were not obvious from the "angle" definition, but which
can be easily seen using components:

-  Distributive property:  $\vec{u}\cdot(\vec{v} + \vec{w}) =
  \vec{u}\cdot\vec{v} + \vec{u}\cdot\vec{w}$

-  Associativity with scalar multiple:  $c(\vec{u}\cdot\vec{v}) =
  (c\vec{u})\cdot\vec{v} = \vec{u}\cdot(c\vec{v})$

    (Note that this property involves 3 different kinds of multiplication:

    -   On the left side, these is a _dot product_ as well as a simple
      _multiplication of two scalars_ $c$ and $\vec{u}\cdot\vec{v}$.

    -   In the middle and on the right, we have the _dot product_ and the _scalar multiple_.

    We can summarize the three products:

    -   A __multiplication of two scalars__ is something you are very familiar
      with.  You take two scalars and get a scalar as a result.

    -   A __scalar multiple of a vector:__ you take a scalar and a vector and
      get a vector as a result.

    -   A __dot product:__  you take two vectors, and get a _scalar_ as a
      result.

    For _positive_ scalar $c$, this property is easy to see from the "angle"
    definition of dot product.  It is a bit trickier for a negative $c$,
    because the angle between the vectors changes, but it can be derived if one
    is careful.  It is very easu to see from the component definition, though.

#### Try it yourself

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/272/setStewart12_3/problem_1a.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/FortLewis/Calc3/13-3-Dot-product/HGM4-13-3-04-The-dot-product.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>

### Angle between vectors

Using the two formulas for dot product, we can calculate the angle between two
vectors that are given using components.

\[\cos\theta = \frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}\]

Then taking the inverse cosine:

\[\left\lvert\theta\right\rvert = \cos^{-1}\left(\frac{\vec{u}\cdot\vec{v}}{\norm{\vec{u}}\norm{\vec{v}}}\right)\]

In 2d, the angle will be positive if it is counterclockwise, and negative if it
is clockwise.  In 3d, the sign will depend on your point of view.  If no
reference point of view is given, the only think that can be determined is the
absolute value of the angle.

#### Example

Find (the absolute value of) the angle between the vectors $\vec{a} =
\v{-2,3,1}$ and $\vec{b} = \v{2,1,-1}$.

We need to calculate the magnitudes and the dot product:

-   $\norm{\vec{a}} = \sqrt{(-2)^2 + 3^2 + 1^2} = \sqrt{14}$
-   $\norm{\vec{b}} = \sqrt{(2)^2 + 1^2 + (-1)^2} = \sqrt{6}$
-   $\vec{a}\cdot\vec{b} = (-2)\cdot 2 + 3\cdot 1 + 1\cdot(-1) = -4 + 3 - 1 = -2$

From the dot product we can see that it is an obtuse angle.
Then 

\[\cos\theta = \frac{\vec{a}\cdot\vec{b}}{\norm{\vec{a}}\norm{\vec{b}}} =
\frac{-2}{\sqrt{14}\sqrt{6}} = -\frac{1}{\sqrt{21}}\]

and

\[\left\lvert\theta\right\rvert = \cos^{-1}\left(-\frac{1}{\sqrt{21}}\right)
\approx 1.79\]

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/angle_example.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Angle between two vectors", src = basename(fname), width=285*2, height=256*2, frameborder="0")
```

#### Try it yourself

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/UMN/calculusStewartET/s_12_3_20.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>

