---
title: Dot Product continued
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Projections

One of the main applications of dot product is in calculating a projection of
one vector onto another

```{r echo=FALSE}
vembedr::embed_youtube("WMUvFjfHUHw")
```

The two important formulas are:

\[\begin{aligned}
\operatorname{comp}_{\vec{b}}\vec{a} &= \frac{\vec{a}\cdot\vec{b}}{\norm{\vec{b}}}\\
\operatorname{proj}_{\vec{b}}\vec{a} &= \frac{\vec{a}\cdot\vec{b}}{\vec{b}\cdot\vec{b}}\vec{b}
\end{aligned}\]


### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart12_3/problem_4.pg",
        width=800, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/NAU/setCalcII/vector_projections.pg",
        width=800, height=500
)
```

## Direction cosines

One way to describe the _direction_ of a vector $\vec{v}$ in 3d is by giving a unit
vector with the same direction as $\vec{v}$.  What is the actual meaning of its
components?

```{r echo=FALSE}
vembedr::embed_youtube("v3Eyi5jINlE")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch4/4-4-MagnitudeAndDirectionDecimals3D.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch4/4-4-MagnitudeAndDirectionFormulas3D.pg",
        width=640, height=500
)
```

