---
title: Vectors
format: html
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}

### Vectors as "Arrows"

```{r}
vembedr::embed_youtube("lR0H7JwK98o")
```

### Addition of 3d vectors

The following illustration shows the addition of two vectors in 3d space

```{r, echo=FALSE}
illustr("../../asymptote/vectors/addition.html", title = "Addition of two vectors", width=285*2, height=256*2)
```

You can interact with the illustration by turning it around and viewing it
from different angles.

### Try it yourself


```{r}
webwork("Library/Hope/Multi1/01-02-Vector-operations/Graphical_vector_ops_04.pg",
width=640, height=500)
```

### Vectors in components

```{r}
vembedr::embed_youtube("IGZKRmgFlhs")
```

Components of a 3d vector:

```{r, echo=FALSE}
illustr("../../asymptote/vectors/components.html", title = "Addition of two vectors", width=285*2, height=256*2)
```

Addition in components:

```{r, echo=FALSE}
illustr("../../asymptote/vectors/addition_boxes.html", title = "Addition of two vectors", width=285*2, height=256*2)
```

### Try it yourself

```{r}
webwork(
"Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch2/2-6-VectorFromOriginToPoint2D.pg",
width=640, height=500)
```

```{r}
webwork(
"Library/Hope/Multi1/01-01-Vectors-in-Rn/Vectors-18a.pg",
width=640, height=500)
```

```{r}
webwork(
"Library/Hope/Multi1/01-01-Vectors-in-Rn/Vectors-02.pg",
width=640, height=500)
```


```{r}
webwork(
"Library/272/setStewart12_2/problem_1a.pg",
width=640, height=500)
```

### Unit vectors and bases

```{r}
vembedr::embed_youtube("Tkki_kHnnmc")
```

Components using bases vectors

```{r, echo=FALSE}
illustr("../../asymptote/vectors/component2.html", title = "Components of a vector", width=285*2, height=256*2)
```

```{r}
webwork(
"Library/272/setStewart12_2/problem_2a.pg",
width=640, height=500)
```


```{r}
webwork(
"Library/Michigan/Chap13Sec1/Q17.pg",
width=640, height=500)
```


