---
title: Vectors
---

### Vectors as "Arrows"

<iframe width="560" height="315" src="https://www.youtube.com/embed/lR0H7JwK98o" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

### Addition of 3d vectors

The following illustration shows the addition of two vectors in 3d space

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/addition.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Addition of two vectors", src = basename(fname), width=285*2, height=256*2, frameborder="0")
```

You can interact with the illustration by turning it around and viewing it
from different angles.

### Try it yourself


<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/Hope/Multi1/01-02-Vector-operations/Graphical_vector_ops_04.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>

### Vectors in components

<iframe width="560" height="315" src="https://www.youtube.com/embed/IGZKRmgFlhs" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Components of a 3d vector:

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/components.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Addition of two vectors", src = basename(fname), width=285*2, height=256*2, frameborder="0")
```

Addition in components:

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/addition_boxes.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Addition of two vectors", src = basename(fname), width=285*2, height=256*2, frameborder="0")
```

### Try it yourself

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/BentleyUniversity/MathematicsOfComputerGraphics/setMoCG-Ch2/2-6-VectorFromOriginToPoint2D.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/Hope/Multi1/01-01-Vectors-in-Rn/Vectors-18a.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/Hope/Multi1/01-01-Vectors-in-Rn/Vectors-02.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/272/setStewart12_2/problem_1a.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>

### Unit vectors and bases

<iframe width="560" height="315" src="https://www.youtube.com/embed/Tkki_kHnnmc" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Components using bases vectors

```{r, echo=FALSE}
fname <- "../../asymptote/vectors/component2.html"
copied <- file.copy(fname, ".", overwrite = TRUE)
htmltools::tags$iframe(title = "Addition of two vectors", src = basename(fname), width=285*2, height=256*2, frameborder="0")
```

<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/272/setStewart12_2/problem_2a.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


<iframe height="500" src="https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=Library/Michigan/Chap13Sec1/Q17.pg
&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple" width="640">
</iframe>


