---
title: Lines and Planes
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Intersection of two planes

If two distinct planes in $\mathbb{R}^3$ are not parallel, they intersect, and the intersection forms a line. A natural question
is to find an equation of that line.


```{r, echo=FALSE}
illustr("../../asymptote/lines_and_planes/planes_intersect.html", title = "An intersection of two planes", width=640, height=420)
```

For an equation of a line, we need a _point on the line_ and its _direction vector_. Since the line lies in both of the planes,
its direction vector must be _orthogonal_ to both of the normal vectors. Let's look at few examples:

### Couple of examples

```{r echo=FALSE}
vembedr::embed_youtube("KBlQu1vMnIA")
```

### Try it yourself

```{r, echo=FALSE}
# Intersection of two planes
webwork(
        "Library/Michigan/Chap17Sec1/Q47.pg",
        width=640, height=500
)
```


## Intersection of a line and a plane

If a line is not _parallel_ to a plane, it will intersect it at a single point. Finding this point is simple: we just need 
to find the time at which the coordinates of a moving object described by the parametric equations of the line satisfy
the equation of the plane:

```{r echo=FALSE}
vembedr::embed_youtube("eRbbMlP3QFY")
```

### Try it yourself

```{r, echo=FALSE}
# Intersection of line and plane
webwork(
        "Library/272/setStewart12_5/problem_14.pg",
        width=640, height=500
)
```

## Equation of a plane containing a given line and a given point.

```{r echo=FALSE}
vembedr::embed_youtube("M1dQrdeTVrA")
```

### Try it yourself

```{r, echo=FALSE}
# Plane by line and point
webwork(
        "Library/UMN/calculusStewartET/s_12_5_34.pg",
        width=640, height=500
)
```

## Equation of a plane _perpendicular to_ a given line, containing a given point.

```{r echo=FALSE}
vembedr::embed_youtube("5X3UMrmKrR4")
```

### Try it yourself

```{r, echo=FALSE}
# Plane perpendicular to a line
webwork(
        "Library/CSUN/Calculus/Lines_planes_2.pg",
        width=640, height=500
)
```

