---
title: Lines and Planes
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}


## Intersection of two planes

If two distinct planes in $\mathbb{R}^3$ are not parallel, they intersect, and the intersection forms a line. A natural question
is to find an equation of that line.


```{r}
illustr("../../asymptote/lines_and_planes/planes_intersect.html", title = "An intersection of two planes", width=640, height=420)
```

For an equation of a line, we need a _point on the line_ and its _direction vector_. Since the line lies in both of the planes,
its direction vector must be _orthogonal_ to both of the normal vectors. Let's look at few examples:

### Couple of examples

```{r}
vembedr::embed_youtube("KBlQu1vMnIA")
```

### Try it yourself

```{r}
# Intersection of two planes
webwork(
        "Library/Michigan/Chap17Sec1/Q47.pg",
        width=640, height=500
)
```


## Intersection of a line and a plane

If a line is not _parallel_ to a plane, it will intersect it at a single point. Finding this point is simple: we just need 
to find the time at which the coordinates of a moving object described by the parametric equations of the line satisfy
the equation of the plane:

```{r}
vembedr::embed_youtube("eRbbMlP3QFY")
```

### Try it yourself

```{r}
# Intersection of line and plane
webwork(
        "Library/272/setStewart12_5/problem_14.pg",
        width=640, height=500
)
```

## Equation of a plane containing a given line and a given point.

```{r}
vembedr::embed_youtube("M1dQrdeTVrA")
```

### Try it yourself

```{r}
# Plane by line and point
webwork(
        "Library/UMN/calculusStewartET/s_12_5_34.pg",
        width=640, height=500
)
```

## Equation of a plane _perpendicular to_ a given line, containing a given point.

```{r}
vembedr::embed_youtube("5X3UMrmKrR4")
```

### Try it yourself

```{r}
# Plane perpendicular to a line
webwork(
        "Library/CSUN/Calculus/Lines_planes_2.pg",
        width=640, height=500
)
```

