---
title: Equations of Planes
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../_common.qmd >}}


## Parametric equations

So far we have learned the following:

*  In $\mathbb{R}^2$, given two _non-parallel_ vectors $\vec{u}$ and $\vec{v}$, any vector can be written as $r\vec{u} + s\vec{v}$
   for some scalars $r$ and $s$.

*  Any point on a line passing through a point $P$ with position vector $\vec{r}_0$ in the direction given
    by a vector $\vec{v}$ in $\mathbb{R}^3$ can be written as $\vec{r}_0 + t\vec{v}$ for some scalar $t$.

*  A line is $\mathbb{R}^3$ can be given by two points $P$ and $Q$.  An equation of such line can be written in _vector form_ as
  $\vec{r} = \vec{r}_0 + t\vec{v}$ where $\vec{r} = \v{x,y,z}$, $\vec{r}_0 = \v{x_0, y_0, z_0}$ is the position vector of $P$,
  and $\vec{v} = \v{v_1,v_2,v_3} = \overrightarrow{PQ}$.

*  We can split the vector equation into components to get _parametric equations_ of the line:
  $$\left\{\begin{aligned}x &= x_0 + tv_1\\y &= y_0 + tv_2\\z &= z_0 + tv_3\end{aligned}\right.$$

Suppose we have to find an equation of the plane containing three points $P$, $Q$ and $R$. We can start by finding
the vectors $\vec{u} = \v{u_1,u_2,u_3} = \overrightarrow{PQ}$ and $\vec{v} = \v{v_1,v_2,v_3} = \overrightarrow{PR}$.
Then any point on the plane can be written as $\vec{r}_0 + r\vec{u} + s\vec{v}$ for some scalars $r$ and $s$, where
$\vec{r}_0 = \v{x,y,z}$ is the position vector of $P$:

```{r}
illustr("../../asymptote/lines_and_planes/plane_param.html", title = "A plane through P, Q and R", width=640, height=420)
```

This will give us a _parametric vector equation_ of the plane: $$\vec{r} = \vec{r}_0 + r\vec{u} + s\vec{v}.$$

This we can split into components to get the _parametric equations_:

$$\left\{\begin{aligned}x &= x_0 + r u_1 + s v_1\\y &= y_0 + r u_2 + s v_2\\z &= z_0 + r u_3 + s v_3\end{aligned}\right.$$

Each pair of values for the scalars $r$ and $s$ corresponds to exactly one point in the plane. We can think about it
as _mapping_ the two-dimensional $rs$-plane onto our plane in $\mathbb{R}^3$ such that the _origin_ maps to $P$,
the point $(1,0)$ maps to $Q$, the point $(0,1)$ maps to $R$, the $r$-axis maps to the line $PQ$, and the $s$-axis maps
to the line $PR$:

```{r}
illustr("../../asymptote/lines_and_planes/plane_param_coordsys.html", title = "An image of the rs-plane", width=640, height=420)
```

### An example

```{r}
vembedr::embed_youtube("N3Y8Xde-BwE")
```

## Implicit equations

There is another way to describe a plane: instead of giving a point and two "direction" vectors, we can instead
give a point and one vector that is _perpendicular_ to the plane.  We call this vector a _normal vector_ of the plane.

Suppose we have a point $P$ with position vector $\vec{r}_0 = \v{x_0, y_0, z_0}$ and a _normal vector_ $\vec{n} = \v{n_1,n_2,n_3}$.
Then for any point $Q(x,y,z)$ with the position vector $\vec{r} = \v{x,y,z}$, the point belongs to the plane if and only if
the vector $\overrightarrow{PQ}$ is _orthogonal_ to $\vec{n}$. Two vectors are orthogonal if and only if their dot product is zero,
which gives us the _implicit vector equation_ of the plane:

$$\vec{n}\cdot (\vec{r} - \vec{r}_0) = 0.$$

Writing this in components and reorganizing will give us the _implicit equation_ of the plane:

$$n_1 x + n_2 y + n_3 z - (n_1 x_0 + n_2 y_0 + n_3 z_0) = 0.$$

```{r}
illustr("../../asymptote/lines_and_planes/plane_implicit.html", title = "Plane defined by a point and a normal vector", width=640, height=420)
```

### An example

```{r}
vembedr::embed_youtube("cdgHV3G9h1k")
```

### Try it yourself

```{r}
webwork(
        "Library/Michigan/Chap17Sec5/Q15.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/Michigan/Chap17Sec5/Q19.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/Michigan/Chap13Sec4/Q13.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/12_Vector_Geometry/12.5_Planes_in_Three-Space/12.5.19.pg",
        width=640, height=500
)
```

### Angle between planes, parallel planes

Two planes are parallel when their normal vectors are scalar multiples of each other.

If two planes are not parallel, they intersect, and the _angle_ between them is the same as
the angle between their normal vectors.


```{r}
vembedr::embed_youtube("FbGGJuE0lHM")
```

### Try it yourself

```{r}
webwork(
        "Library/UMN/calculusStewartET/s_12_5_50.pg",
        width=640, height=500
)
```

```{r}
webwork(
        "Library/Rochester/setVectors4PlanesLines/ur_vc_2_18.pg",
        width=640, height=500
)
```
