---
title: Absolute Maxima and Minima
format: html
layout: article
reference-location: margin
knitr:
  opts_chunk: 
    echo: false
---

{{< include ../../_common.qmd >}}

```{r, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
```

## Functions of One Variable

In Calc I you have learned about _absolute maxima and minima_ of function. Two principles were
important for finding those:

1.  A _continuous function_ on a _closed interval_ has both absolute maximum and absolute minimum.

2.  An absolute extrema (a maximum or minimum) on a closed interval can be attained either
   at a _critical point_ inside the interval, or at an _end point_ of the interval.

To find the absolute maximum or minimum of a _continuous function_ on a _closed interval_, one
has to find all the critical points _inside_ the interval, and plug all those points _and the endpoints_
or the interval to the function, to see at which of the points the function will have the highest
or the lowest value.

## Functions of several variables

The same principles apply in several variables:

Given a set $A$ of points in $\mathbb{R}^n$:

*   A point $p \in \mathbb{R}^n$ is a _boundary point_ of a set $A \subseteq \mathbb{R}^n$ if
    any ball centered at $p$ contains points that are in $A$ as well as points that
    are not in $A$. A point that belongs to $A$ but is not a boundary point is called an
    _interior point_ of $A$.  A point $p \in A$ is an _interior point_ of $A$ if there
    exists a ball $B$ centered at $p$ such that $B \subseteq A$.

*   A set $A \subseteq \mathbb{R}^n$ is __closed__ if it contains _all of its boundary points_.

*   A set $A \subseteq \mathbb{R}^n$ is __open__ if it contains _none of its boundary points_.

### Examples

Take for example the sets $G = \left\{(x,y) \in \mathbb{R}^n \mid x^2 + y^2 < 1\right\}$ and 
$F = \left\{(x,y) \in \mathbb{R}^n \mid x^2 + y^2 \le 1\right\}$. The boundary of each of the set
is the _unit circle_ $\left\{(x,y) \in \mathbb{R}^2 \mid x^2 + y^2 = 1\right\}$.
The set $F$ contains all the boundary points, so it is _closed_, while the set $G$ does not contain any
of the boundary points, so it is _open_.

### The two principles

1.  A _continuous function_ on a _bounded closed set_ in $\mathbb{R}^n$ has both absolute maximum and absolute minimum.

2.  An absolute extrema (a maximum or minimum) of a continuous function on a bounded closed set in $\mathbb{R}^n$
   can be attained either at a _critical point_ that is an interior point of the set, or at a _boundary point_ of the set.


### The procedure

Given a continuous function $f:\mathbb{R}^n \to \mathbb{R}$ defined on a bounded closed set $A \subseteq \mathbb{R}^n$, the task is to find
points in $A$ where $f$ has absolute maximum or absolute minimum. We do it by first finding all the _candidate points_
where an absolute maximum or minimum are possible, then plugging all those points into the function to see which of them will give the highest or lowest value.

The set of _candidate points_ will consist of

1.  Critical points in the interior of the set
2.  _Candidate points_ on the boundary.

Finding candidate points on the boundary can generally be very hard, but if the set $A$ is nice and the boundary consists of one or more
parts that can be _parametrized_ using $n-1$ variables, then it will be reduced to $n-1$ dimensional version of the same problem.  The
following video shows an example for $n=3$, with the set $A$ being a cube.  The example is very long, because the boundary of the three-dimensional cube consists of
6 squares, and each square in turn has 4 boundary segments. Altogether, critical points must be found inside one cube, six squares, and twelve
intervals, and the intervals have eight endpoints among them.  You may want to watch it with higher speed, or fast forward through some of the calculations.

```{r echo=FALSE}
vembedr::embed_youtube("AH16GJQAnNE")
```

The following illustration shows the cube with all the candidate points for maxima or minima: the critical point inside the cube
(<span style="color:red;">red</span>),
the six critical points inside the sides
(<span style="color:green;">green</span>),
the twelve critical points on the edges
(<span style="color:magenta;">magenta</span>),
and the eight corners
(<span style="color:blue;">blue</span>):

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/extrema/cube.html", title = "A cube from the video example", width=500, height=500)
```

### Simpler examples

The following two examples show finding absolute maxima and minima of two variable functions.

```{r echo=FALSE}
vembedr::embed_youtube("umVQepsFykk")
```

The graph of the function looks like this:

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/extrema/Max_Min-6.html", title = "An example on a square", width=500, height=500)
```

```{r echo=FALSE}
vembedr::embed_youtube("hAgD6Ef0GvQ")
```

The graph of the function looks like this:

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/extrema/Max_Min-7.html", title = "An example on a disk", width=500, height=500)
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_7/problem_12.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_7/problem_14.pg",
        width=640, height=500
)
```

