---
title: Lagrange Multipliers
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
```


```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Method of Lagrange Multipliers

Earlier, as a part of a problem were we were finding absolute maxima and minima of a continuous function on a closed disk,
we had to find maxima and minima of the function $f(x,y) = 2x^3 - y^4$ on the boundary of the unit disk $x^2 + y^2 \le 1$.
The boundary has the equation $x^2 + y^2 = 1$.  We did it by splitting the boundary into the upper and lower semicircles, parametrizing
each of them by $x$, and solving the two resulting single variable optimization problems. We ended up with critical points at $(0, \pm 1)$,
and $(-\frac{1}{2}, \pm\frac{\sqrt{3}}{2})$, and with end-points $(\pm 1, 0)$.  The absolute maximum was attained at $(1,0)$, the absolute minimum at $(-1,0)$.

The graph of the function looks like this:

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/extrema/Max_Min-7.html", title = "An example on a disk", width=500, height=500)
```

Now we are going to learn a different way of finding the maximum and minimum of $f$ on the boundary.  This kind of problem is called 
_constrained optimization_: find the maximum and minimum of a function of several variables $f(x_1, x_2, \dots, x_n)$ subject to
a constraint given by an equation in the form $g(x_1, x_2, \dots, x_n) = c$ for some function $g$ and a constant $c$.


```{r echo=FALSE}
vembedr::embed_youtube("Qa7RlLacbtQ")
```

The contour plot of the function $f(x,y) = 2x^3 - y^4$ together with the unit circle $x^2 + y^2 = 1$ is shown below.
The points where a contour is _tangent_ to the unit circle are marked and labeled.

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/extrema/Max_Min-8.html", title = "A contour plot of the function f with the unit circle.", width=500, height=500)
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap15Sec3/Q05.pg",
        width=640, height=500
)
```

## More examples

### Ellipse constraint

Find the absolute maxima and minima of $f(x,y) = 3x^2 + 2xy$ on the ellipse $4x^2 + y^2 = 5$.

```{r echo=FALSE}
vembedr::embed_youtube("uXjWBD4gZNw")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Rochester/setVmultivariable7MaxMin/ur_vc_7_13.pg",
        width=640, height=500
)
```

### Unbounded constraint

Find the absolute maxima and minima of $f(x,y) = 3x^2 + 2y^2$ on the hyperbola $xy = 6$.

```{r echo=FALSE}
vembedr::embed_youtube("cwoUQlnN3Kc")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/UMN/calculusStewartET/s_14_8_3.pg",
        width=640, height=500
)
```

### Three variables

Find the absolute maxima and minima of $f(x,y) = 3x^2 + 2y^2$ on the hyperbola $xy = 6$.

```{r echo=FALSE}
vembedr::embed_youtube("w1IUl5cpJq0")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_8/problem_3.pg",
        width=640, height=500
)
```

