---
title: Limits of Functions of Several Variables
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Directional derivatives

The derivative of a function of one variable will give us the _rate of change_
of that function:  it will tell us _how many times faster is the output of the
function changing compared to the input_.

When attempting to extend this to functions of several variables, we face the
following problem: the way the output changes will depend not only on the
magnitude of the input change, but also on its direction.

If we want to find the rate of change of a function of several variables, we
need to specify the direction, which can be done by specifying a _unit vector_.

### Definition

Given a function $f$ of $n$ variables ($f:\mathbb{R}^n \to \mathbb{R}$), a
point $A(a_1, a_2, \dots, a_n)$, and a _unit vector_ $\vec{u}$, we define the
__directional derivative__ of $f$ at $A$ in the direction $\vec{u}$ the
following way:

$$f_{\vec{u}}(a_1, a_2, \dots, a_n) = \lim_{h \to 0} 
\frac{f(x_1 + hu_1, x_2 + hu_2, \dots, x_n+hu_n) - f(x_1, x_2, \dots,
x_n)}{h}$$

### Another way to think about it

Suppose we have a line passing through the point $A$ with the velocity vector
$\vec{u}$ (since this is a unit vector, the speed of the motion is 1).  The
equation of the line will be

$$\left\{
\begin{aligned}
x_1(t) &= a_1 + u_1 t\\
x_2(t) &= a_2 + u_2 t\\
&\vdots\\
x_n(t) &= a_n + u_n t
\end{aligned}
\right.$$

If we substitute the $x_i$s into the function $f$, we will get a real function
of single variable $t$:

$$\begin{aligned}
g(t) &= f\left(x_1(t), x_2(t), x_3(t), \dots, x_n(t)\right)\\
&= f\left(a_1 + u_1 t, a_2 + u_2 t, a_3 + u_3 t, \dots, a_n + u_n t\right)
\end{aligned}$$

The derivative of $g$ at $0$ is then the _directional derivative_ of $f$ at $A$
in the direction $\vec{u}$.

```{r echo=FALSE}
vembedr::embed_youtube("HiF3AjuYXoA")
```

## Partial derivatives

Let's try to calculate a directional derivative of a function $f$ of two
variables in the direction of the _unit basis vector_ $\vec{i} = \v{1,0}$:

$$f_{\vec{i}}(x,y) = \lim_{h\to 0} \frac{f(x+1h, y + 0h) - f(x,y)}{h}$$

This can be interpreted as a derivative of the function $f(x,y)$ where the
variable $y$ is __fixed__ and only the variable $x$ changes.

### Partial evaluation and partial functions

Given a function $f:\mathbb{R}^n \to \mathbb{R}$, we can plug in a number for
one of the variables $x_1$, $x_2$, ..., $x_n$.  The remaining $n-1$ variables
will still be undetermined, so we will end up with a function that has $n-1$
variables.  We can repeat this several times, in the end plugging in $k$
numbers for $k$ of the variables, and ending up with a function of $n-k$
variables.

This process is called _partial evaluation_, and the resulting function a
_partial function_ of $f$.

In particular, we can plug in values for _all but one_ of the variables, ending
with a _partial function_ of one variable.  We can then take a derivative of
this function.

### Partial derivatives definition:

Given a function $f:\mathbb{R}^n \to \mathbb{R}$, then for every index $i =
1,2,\dots,n$ we can define the _partial derivative_ with respect to $x_i$:

We will consider all the variables except $x_i$ to be fixed numbers, which will
turn $f$ into a partial function with only one variable, $x_i$.  Then the
_partial derivative of $f$ with respect to $x_i$_ is the derivative of this
partial function.  It is also a _directional derivative_ of $f$ in the
direction of the $i$-th standard basis vector.

We use the notation $f_{x_i}(x_1, x_2, \dots, x_n)$ or, equivalently,

$$\frac{\partial f}{\partial x_i} (x_1, x_2, \dots, x_n)$$

#### For $n = 2$ we get two partial derivatives:

-  Partial derivative with respect to $x$:

   $$f_x(x,y) = \frac{\partial f}{\partial x}(x,y) = f_{\vec{i}}(x,y) =
   \lim_{h\to 0}\frac{f(x+h,y) - f(x,y)}{h}$$

-  Partial derivative with respect to $y$:

   $$f_y(x,y) = \frac{\partial f}{\partial y}(x,y) = f_{\vec{j}}(x,y) =
   \lim_{h\to 0}\frac{f(x,y+h) - f(x,y)}{h}$$

#### For $n = 3$ we get three partial derivatives:

-  Partial derivative with respect to $x$:

   $$f_x(x,y,z) = \frac{\partial f}{\partial x}(x,y,z) = f_{\vec{i}}(x,y,z) =
   \lim_{h\to 0}\frac{f(x+h,y,z) - f(x,y,z)}{h}$$

-  Partial derivative with respect to $y$:

   $$f_y(x,y,z) = \frac{\partial f}{\partial y}(x,y,z) = f_{\vec{j}}(x,y,z) =
   \lim_{h\to 0}\frac{f(x,y+h,z) - f(x,y,z)}{h}$$

-  Partial derivative with respect to $z$:

   $$f_z(x,y,z) = \frac{\partial f}{\partial z}(x,y,z) = f_{\vec{k}}(x,y,z) =
   \lim_{h\to 0}\frac{f(x,y,z+h) - f(x,y,z)}{h}$$


```{r, echo=FALSE}
illustr("../../../asymptote/multivar/derivatives/partials.html", title = "Graph showing partial derivatives of a function of two variables.", width=500, height=500)
```

### Calculating partial derivatives

```{r echo=FALSE}
vembedr::embed_youtube("VkAGPgLntnM")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/14_Differentiation_in_Several_Variables/14.3_Partial_Derivatives/14.3.42.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap14Sec2/Q08.pg",
        width=640, height=500
)
```

## Differentiability and tangent plane

```{r echo=FALSE}
vembedr::embed_youtube("uYI7VFMk0Bg")
```

The following graph shows the tangent plane to a surface, together with traces
that are graphs of the _partial functions_.

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/derivatives/tangent.html", title = "Tangent plane to a surface at a point", width=500, height=500)
```

## A non-differentiable function

The existence of _partial derivatives_ of $f$ at a point $A$ does not guarantee that $f$ is differentiable at $A$:

```{r echo=FALSE}
vembedr::embed_youtube("WAtAvC9DQqE")
```

The following graph shows the function $f(x,y) = \frac{xy}{x^2 + y^2}$ with a
"tangent plane" at $(0,0)$ which is not actually tangent to the surface:

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/derivatives/nodiff.html", title = "Graph of a non-differentiable function at (0,0)", width=500, height=500)
```

## Directional derivatives of a differentiable function

```{r echo=FALSE}
vembedr::embed_youtube("wL7_o8ZI2LM")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/14_Differentiation_in_Several_Variables/14.5_The_Gradient_and_Directional_Derivatives/14.5.25.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap14Sec5/Q19.pg",
        width=640, height=500
)
```
