---
title: Gradient vector, higher derivatives, ...
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Review

For a function of several variables, we have defined the _directional
derivative_ in the direction of a _unit vector_ $\vec{u}$ as

$$f_{\vec{u}}(a_1, a_2, \dots, a_n) = \lim_{h \to 0} 
\frac{f(x_1 + hu_1, x_2 + hu_2, \dots, x_n+hu_n) - f(x_1, x_2, \dots,
x_n)}{h}$$

We then defined _partial derivatives_ of $f$ as directional derivatives in the
direction of one of the _standard basis vectors_, that is unit vectors in the
directions of the coordinate axes.

In two dimensions, these are the vectors $\vec{i}$ and $\vec{j}$, leading to
the partial derivative _with respect to $x$_:

$$\frac{\partial f}{\partial x}(x,y) = f_x(x,y) = f_{\vec{i}}(x,y) = D_{\vec{i}} f(x,y)$$

and 

$$\frac{\partial f}{\partial y}(x,y) = f_y(x,y) = f_{\vec{j}}(x,y) = D_{\vec{j}} f(x,y)$$

In three dimensions we also add the partial derivative with respect to $z$,
corresponding to the vector $\vec{k}$.

We also learned what it means for a function of several variables to be
_differentiable_:  It is not enough to just have all the partial derivatives.
A function $f:\mathbb{R}^n \to \mathbb{R}$ is _differentiable at a point $A =
(a_1, a_2, \dots, a_n)$_ if it can be well approximated near $A$ by a linear
function $L:\mathbb{R}^n \to \mathbb{R}$.  Formally. $f$ is differentiable at
$A$ if there exists a linear function $L$ such that 

$$\lim_{(x_1,x_2,\dots, x_n)\to (a_1,a_2,\dots,a_n)}\frac{f(x_1,x_2,\dots,x_n) -
L(x_1,x_2,\dots,x_n)}{\norm{\v{x_1,x_2,\dots,x_n} - \v{a_1,a_2,\dots,a_n}}} =
0$$

In two variables, it simply means that the surface graph of the function has a
_tangent plane_.

For example, the following surface graph shows a function that has _all
directional derivatives_ at the origin, but is not differentiable there:

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/derivatives/nodiff1.html", title = "A surface graph of a function that has all directional derivatives at (0,0) but is nit differentiable there", width=500, height=500)
```

Finally, we learned that for a _differentiable_ function, the __gradient
vector__ defined as

$$\vec{\nabla}f(x_1,x_2,\dots,x_n) = \v{
\frac{\partial f}{\partial x_1}(x_1, x_2,\dots,x_n),
\frac{\partial f}{\partial x_2}(x_1, x_2,\dots,x_n),
\dots,
\frac{\partial f}{\partial x_n}(x_1, x_2,\dots,x_n)
}$$

is very useful.  For example, _if $f$ is a differentiable function at $A$_,
then for any unit vector $\vec{u}$ the _directional derivative_ of $f$ at $A$
in the direction $\vec{u}$ can be calculated as

$$D_\vec{u} f(a_1,a_2,\dots,a_n) = \vec{\nabla}f(a_1,a_2,\dots,a_n)\cdot
\vec{u}.$$

The next video talks a bit more about the gradient vector:

```{r echo=FALSE}
vembedr::embed_youtube("7ecWqIYcvB0")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_6/problem_9.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/14-4-Gradients-in-plane/HGM4-14-4-34-Gradients-etc.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap17Sec1/Q55.pg",
        width=640, height=500
)
```

## The Chain Rule

```{r echo=FALSE}
vembedr::embed_youtube("l8twSna0AO0")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap14Sec6/Q06.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/14_Differentiation_in_Several_Variables/14.6_The_Chain_Rule/14.6.13.pg",
        width=640, height=500
)
```

## Higher order derivatives

```{r echo=FALSE}
vembedr::embed_youtube("Sby_4al4zzE")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_3/UR_VC_5_14.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/14_Differentiation_in_Several_Variables/14.3_Partial_Derivatives/14.3.67.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap14Sec7/Q15.pg",
        width=640, height=500
)
```
