---
title: Functions of Several Variables
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(tidyverse)
```


```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Functions of Several Variables

A (real) function of $n$ variables is a function $f:\mathbb{R}^n \to \mathbb{R}$. It's domain is a subset of $\mathbb{R}^n$, and we can think
about it as a way to assign each point in the domain a real number.

```{r echo=FALSE}
vembedr::embed_youtube("lh3Qau8NSns")
```

### Examples of domains

*  _Find the domain of the function $f(x,y,z) = \ln(36 - x^2 - y^2 - z^2)$:_

   Here we must have $36 - x^2 - y^2 - z^2 > 0$, or $x^2 + y^2 + z^2 < 36$ which describes the _interior_ of the sphere with center at
   the origin and with radius 6.  Another name for this is an _open ball_ with center at the origin and radius 6.

*  _Find the domain of the function $g(x,y) = \sqrt{y^2 - x^2}$:_

   Here we must have $y^2 - x^2 \ge 0$, or $y^2 \ge x^2$. Since both sides of the inequality are positive and square root is
   an _increasing_ function, we can take square root of both sides and get $\left\lvert y\right\rvert \ge \left\lvert x\right\rvert$
   which describes the part of the plane at and above the graph of $y = \left\lvert x\right\rvert$ or at and below the graph of
   $y = -\left\lvert x\right\rvert$.

    ```{r, echo=FALSE}
    illustr("../../../asymptote/multivar/intro/2dcone.html", title = "Domain of g", width=300, height=300)
    ```

*  _Find the domain of the function $h(x,y,z) = \sqrt{z^2 - x^2 - y^2}$:_

   Here we must have $z^2 - x^2 - y^2 \ge 0$, or $z^2 \ge x^2 + y^2$. Since both sides of the inequality are positive and square root is
   an _increasing_ function, we can take square root of both sides and get $\left\lvert y\right\rvert \ge \sqrt{x^2 + y^2}$
   which describes the part of the space at and inside the cone $y = \pm\sqrt{x^2 + y^2}$.

    ```{r, echo=FALSE}
    illustr("../../../asymptote/multivar/intro/cone.html", title = "Domain of g", width=500, height=500)
    ```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/FortLewis/Calc3/12-1-Two-variable-functions/HGM4-12-1-22-Functions-of-two-variables.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Union/setMVfunctions/an14_1_1.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/WHFreeman/Rogawski_Calculus_Early_Transcendentals_Second_Edition/14_Differentiation_in_Several_Variables/14.1_Functions_of_Two_or_More_Variables/14.1.7.pg",
        width=640, height=900
)
```
