---
title: Limits of Functions of Several Variables
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\v}[1]{\left\langle #1\right\rangle}
\newcommand{\norm}[1]{\left\lvert#1 \right\rvert}

```{r, echo=FALSE, messages=FALSE, errors=FALSE, warnings=FALSE}
library(tidyverse)
```


```{r, echo=FALSE}
webwork <- function(path, width, height)
{
    htmltools::tags$iframe(title = "Exercise", 
                           src=paste0("https://webwork.svsu.edu/webwork2/html2xml?
&answersSubmitted=0
&sourceFilePath=", path, "&problemSeed=123
&displayMode=MathJax
&courseID=daemon_course
&userID=daemon
&course_password=daemon
&outputformat=simple"), width=width, height=height, frameborder="0")
}

illustr <- function(fname, title, width, height)
{
    copied <- file.copy(fname, ".", overwrite = TRUE)
    htmltools::tags$iframe(title = title, src = basename(fname), width=width, height=height, frameborder="0")
}
```

## Definition of limit

Given a function $f$ of $n$ variables ($f:\mathbb{R}^n \to \mathbb{R}$), and a
point $A(a_1, a_2, \dots, a_n)$, we say that the limit of $f(x_1, x_2, \dots,
x_n)$ at $(a_1, a_2, \dots, a_n)$ is a real number $L$, or symbolically

$$\lim_{(x_1,x_2,\dots,x_n)\to(a_1,a_2,\dots,a_n)} f(x_1, x_2, \dots, x_n) = L$$

if for every $\varepsilon > 0$ there exists a $\delta > 0$ such that
$\norm{f(x_1,x_2,\dots,x_n) - L}< \varepsilon$ whenever $0 < \norm{\v{x_1,
x_2, \dots, x_n} - \v{a_1, a_2, \dots, a_n}}< \delta$.

```{r echo=FALSE}
vembedr::embed_youtube("pCbejhh19IQ")
```

## Examples

Some examples of functions that don't have a limit at some point.

### Example 1

```{r echo=FALSE}
vembedr::embed_youtube("3JT5dabAxLA")
```

The surface graph of the function $\displaystyle f(x,y) = \frac{xy}{x^2+y^2}$ from the previous video is

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/limits/nolimit_1_surface.html", title = "Graph of a function with no limit at (0,0).", width=300, height=300)
```

### Example 2

```{r echo=FALSE}
vembedr::embed_youtube("6dB7dlxuFc8")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap12Sec6/Q13.pg",
        width=640, height=500
)
```

```{r, echo=FALSE}
webwork(
        "Library/Michigan/Chap12Sec6/Q15.pg",
        width=640, height=500
)
```

## Some limits that exist

If a limit exists, we have number of ways to find it and prove that the number
we found is really the limit.

### Squeeze Theorem

```{r echo=FALSE}
vembedr::embed_youtube("IqxapaLna6E")
```

The surface graph of the function $\displaystyle f(x,y) = \frac{xy}{\sqrt{x^2+y^2}}$ from the previous video is

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/limits/limit_squeeze.html", title = "Graph of a function squeezed between |y| and -|y|.", width=300, height=300)
```

The graph shows the function "squeezed" between $\left\lvert y\right\rvert$ and
$-\left\lvert y\right\rvert$.

### Replacement Theorem

```{r echo=FALSE}
vembedr::embed_youtube("97qP_LcHUcA")
```

The surface graph of the function $\displaystyle f(x,y) = \frac{x^3 - y^3}{x^2 + xy + y^2}$ from the previous video is

```{r, echo=FALSE}
illustr("../../../asymptote/multivar/limits/limit_replacement.html", title = "Graph of a function with limit at (0,0).", width=300, height=300)
```

You can see that it is really just a linear function, except that it is
undefined at $(0,0)$ (you cannot see that from the graph, of course).

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_2/problem_3.pg",
        width=640, height=500
)
```

## Continuity

```{r echo=FALSE}
vembedr::embed_youtube("xHEfZ0ju71c")
```

### Try it yourself

```{r, echo=FALSE}
webwork(
        "Library/272/setStewart14_2/UR_VC_5_3.pg",
        width=640, height=500
)
```

